### 2024-12-09 Draft_Cleaning (1).Rmd
copied existing data over

---
title: "DRAFT_Cleaning"
output: html_document
date: "2024-12-04"
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggformula)

dredge_data<-read_csv("DredgeData.csv")
survey_data<-read_csv("SurveyData.csv")
gage_data<-read_csv("UMR_IWW_1999_2021.csv")

gage_data2<- gage_data |>
  group_by(Date)|> 
   summarize(across(everything(),~first(na.omit(.)), .names = "{.col}"))|>
  mutate(DATE_START= as_date(dmy(Date)))|>
    arrange(DATE_START) 
gage_data2<-gage_data2|>
    mutate(across(c(-Date,-DATE_START), 
                ~ ifelse(is.na(.), 
                         coalesce((lag(.) + lead(.)) / 2, lag(.), lead(.)), 
                         .)))|>
  select(-Date)

survey_Data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         SURVEYJOBIDPK == str_replace_all(SURVEYJOBIDPK, "IL", "IW"),
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0, 
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,UP_RIV_MIL,DN_RIV_MIL)

dredge_data_cleaning <- dredge_data |>
   filter(
          DREDGINGPURPOSE != "harbor",
           EXECYEAR <= 2022)|>
  mutate(
    river_code = str_sub(PREHYDROSURVEY_ID, 1, 2),
    pool_code = str_sub(PREHYDROSURVEY_ID, 4, 5),
    district_code = str_sub(PREHYDROSURVEY_ID, 7, 9),
    river_mile = str_sub(PREHYDROSURVEY_ID, 23, 30), 
    year_iteration = str_sub(PREHYDROSURVEY_ID, 32,34),
    SURVEYJOBIDPK = ifelse(
      str_detect(PREHYDROSURVEY_ID, "MVRUM|MVRIW"), 
      PREHYDROSURVEY_ID, 
      paste0(district_code, river_code, pool_code, "_", river_mile, "_", year_iteration)  
    ),
      SURVEYJOBIDPK = str_replace_all(SURVEYJOBIDPK, "IL", "IW"), 
     DATE_START = as_date(mdy_hm(DATE_START)),
    SDSFEATURENAME = FEATURENAME)|>
  select(SURVEYJOBIDPK,EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,DATE_START, UP_RIV_MIL,DN_RIV_MIL)

combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") & 
         DATE_START <= as.Date("2021-09-30"))

combined_data <-combined_data |>
  mutate(lead_7 = (DATE_START - 7),
         lead_14 = (DATE_START - 14))
Nas<- combined_data|>
  filter(if_any(everything(),is.na))

gage_survey <- combined_data |>
  left_join(gage_data2, by = "DATE_START")

na<-gage_survey |>
  select(everything()) |>
  summarise_all(list(~sum(is.na(.))))
na

NA_by_row<- gage_survey |>
  is.na() |>
  rowSums()

gf_histogram(~NA_by_row)

gage_survey_filt <- gage_survey |>
  filter(complete.cases(gage_survey))

gage_survey_full <- gage_survey_filt|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

Nas<- gage_survey_full|>
  select(-lead_7,-lead_14)|>
  filter(if_any(everything(),is.na))

gage_survey_cleaned <- gage_survey_full |>
  filter(complete.cases(gage_survey_full))

final_na<-gage_survey_cleaned|>
 select(-lead_7,-lead_14)|>
  filter(if_any(everything(),is.na))

### Create factored season variable that could be used in analysis
date_to_season <- function(date) {
  month <- as.numeric(format(date, "%m"))
  day <- as.numeric(format(date, "%d"))
  # Define the seasons based on date ranges
  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
    return("Spring")
  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
    return("Summer")
  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
    return("Fall")
  } else {
    return("Winter")
  }
}

gage_survey_cleaned <- gage_survey_cleaned |>
  mutate("Season" = sapply(DATE_START,date_to_season))

```

```{r}
gage_survey_PCA <- gage_survey_cleaned |>
select(where(is.numeric), -EXECYEAR)

dredge_numeric<-gage_survey_cleaned|>
   select(where(is.numeric), -EXECYEAR)
correlations <-cor(dredge_numeric, use = "pairwise.complete.obs")
corrplot::corrplot(correlations)

dredge_prcomp <- gage_survey_PCA|>
  select(-VOLUMEDREDGED)

dredge_PCA = prcomp(dredge_prcomp, scale = T)
biplot(dredge_PCA)

variable_contributions<- dredge_PCA$rotation
head(variable_contributions)

set.seed(12)
ctrl = trainControl(method = "cv", number = 5)

fit_pca = train(VOLUMEDREDGED ~ ., 
            data = gage_survey_PCA,
            method = "glm",
            preProc = c("pca"),
            trControl = ctrl)
fit_pca

summary(fit_pca$finalModel)
varImp(fit_pca)
variable_contributions = fit_pca$preProcess$rotation
head(variable_contributions)

data.frame(variable_contributions) |>
  gf_point(PC2 ~ PC1)

data.frame(variable_contributions) |>
  gf_text(PC2 ~ PC1, 
          label = row.names(variable_contributions))

predictor_matrix <- gage_survey_PCA |>
  dplyr::select(-VOLUMEDREDGED) %>% # Remove the response variable
  as.matrix() %>%
  scale() #centers and scales

# The symbol %*% does matrix multiplication
data_in_PC_space = predictor_matrix %*% variable_contributions

data.frame(data_in_PC_space) %>%
  gf_point(PC2 ~ PC1)

```

###xgBoost
```{r}
library(xgboost)
gage_xgb <- gage_survey |>
  select(-Date_obvs,-Date_7,-Date_14,-DATE_START,
         -lead_7,-lead_14,-SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE)|>
    mutate_if(is.character, factor)

ctrl = trainControl(method = "cv", number = 10) # Increase this to 10 to get more consistent RMSEs
fit_dredge = train(VOLUMEDREDGED ~ ., 
                 data = gage_xgb,
                 method = "xgbTree",
                 tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:4, 
                                        eta = 0.3, gamma = 0, 
                                        colsample_bytree = .8, 
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0, 
                 trControl = ctrl, 
                 na.action = na.exclude)

varImp(fit_dredge)

import_matrix = xgb.importance(model = fit_dredge$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
###Cross Validation of Model
```{r}
set.seed(123)
n = nrow(gage_xgb)
n_outer_folds = 10

cv_pred = vector(length = n)

groups = rep(1:n_outer_folds, length = n)
cv_groups = sample(groups, n)

for(ii in 1:n_outer_folds){
  in_test = (cv_groups == ii)
  in_train = (cv_groups != ii)

  ctrl = trainControl(method = "cv", number = 5)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ ., 
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:6,
                                        eta = 0.3, gamma = 0, 
                                        colsample_bytree = 1, 
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0, 
                       trControl = ctrl,
                       na.action = na.exclude)

  # "Frankenstein" vector of predictions
  cv_pred[in_test] = predict(fit_dredge_DCV, 
                             newdata = gage_xgb[in_test, ])

}

fit_dredge_DCV

confusion_matrix <- confusionMatrix(cv_pred[in_test], gage_xgb[in_test, ]$VOLUMEDREDGED)
print(confusion_matrix)

# I'll use RMSE to assess the model, because that's the metric I used to select the model.
# The model with the best RMSE may or may not be the one with the best MAE.
# RMSE
sqrt(mean((cv_pred - gage_xgb$VOLUMEDREDGED)^2)) # This is the most honest 
                                         # assessment of accuracy

dredge_data_cleaning|>
  group_by(POOL)|>
  summarize(count())
```

### 2024-12-09 Draft_Cleaning.Rmd
copied existing data over

--- 2024-12-09 Draft_Cleaning (1).Rmd
@@
 library(caret)
+library(ggfortify)
 library(ggformula)
@@
     arrange(DATE_START)
+
 gage_data2<-gage_data2|>
@@
 dredge_PCA = prcomp(dredge_prcomp, scale = T)
+
 biplot(dredge_PCA)
@@
 gage_xgb <- gage_survey |>
-  select(-Date_obvs,-Date_7,-Date_14,-DATE_START,
-         -lead_7,-lead_14,-SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
+  select(-lead_7, -lead_14,-DATE_START,
+         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
          -DREDGINGPURPOSE)|>
@@
 
-ctrl = trainControl(method = "cv", number = 10) # Increase this to 10 to get more consistent RMSEs
+ctrl = trainControl(method = "cv", number = 10) 
 fit_dredge = train(VOLUMEDREDGED ~ .,
@@
                  method = "xgbTree",
-                 tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:4, 
+                 tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
                                         eta = 0.3, gamma = 0,
@@
 varImp(fit_dredge)
+
+fit_dredge
 
@@
 
-  ctrl = trainControl(method = "cv", number = 5)
+  ctrl = trainControl(method = "cv", number = 10)
   fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
@@
                        method = "xgbTree",
-                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:6,
+                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
                                         eta = 0.3, gamma = 0,
@@
 
-fit_dredge_DCV
+tune_grid <- expand.grid(
+  nrounds = c(50, 100, 150), 
+  max_depth = c(3, 6, 9), 
+  eta = c(0.1, 0.2, 0.3), 
+  gamma = c(0, 0.1), 
+  colsample_bytree = c(0.6, 0.8), 
+  min_child_weight = c(1, 2, 5), 
+  subsample = c(0.8, 1)
+)
 
-confusion_matrix <- confusionMatrix(cv_pred[in_test], gage_xgb[in_test, ]$VOLUMEDREDGED)
+set.seed(123)
+ctrl = trainControl(method = "cv", number = 10, savePredictions = "final")
+
+fit_dredge2 = train(VOLUMEDREDGED ~ .,
+                   data = gage_xgb,
+                   method = "xgbTree",
+                   tuneGrid = tune_grid,
+                   verbosity = 0,
+                   trControl = ctrl,
+                   na.action = na.exclude)
+
+# Model performance
+fit_dredge2
+
+# Model evaluation
+predictions <- fit_dredge2$pred  # Predictions from cross-validation
+head(predictions)
+
+# RMSE, R-Squared, etc.
+cv_rmse <- sqrt(mean((predictions$obs - predictions$pred)^2))
+cv_r2 <- cor(predictions$obs, predictions$pred)^2
+
+cat("RMSE: ", cv_rmse, "\n")
+cat("R-Squared: ", cv_r2, "\n")
+
+##Og
+train<-gage_xgb[in_train, ]
+best_tune<-fit_dredge_DCV$results
+best_tune
+confusion_matrix <- confusionMatrix(cv_pred[in_test], gage_xgb[in_test, ])
 print(confusion_matrix)

### 2024-12-09 Lesson 11 examples--PCA (1).Rmd
copied existing data over

---
title: "Lesson 12 examples"
author: "Abra Brisbin"
date: "11/15/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(caret)
library(MASS) # For simulating correlated data; load this before dplyr to avoid masking `select`

library(dplyr)
library(ggformula)
library(readr)

library(corrplot)
library(RColorBrewer)

```

## PCA

```{r}

data("segmentationData")

```
```{r}
head(segmentationData)
```

```{r}

segmentationData2 <- segmentationData %>%
  dplyr::select(-c(Cell, Case))

head(segmentationData2)

```

```{r}

seg_numeric = select_if(segmentationData2, is.numeric)
correlations <-cor(seg_numeric, use = "pairwise.complete.obs")

# Here, I'm specifying that I want the version of the `corrplot` function that's in the `corrplot` package, to avoid confusion with a function of the same name in the `pls` package.
corrplot::corrplot(correlations, type="upper", order="hclust",
         col=rev(brewer.pal(n=8, name="RdYlBu")))

```

## No PCA
```{r}

set.seed(12)
ctrl = trainControl(method = "cv", number = 5)
fit = train(Class ~ ., 
            data = segmentationData2,
            method = "glm",
            family = "binomial",
            trControl = ctrl)
fit

```

```{r}
summary(fit$finalModel)
```

PCA
```{r}

set.seed(12)
ctrl = trainControl(method = "cv", number = 5)

fit_pca = train(Class ~ ., 
            data = segmentationData2,
            method = "glm",
            family = "binomial",
            preProc = c("pca"),
            trControl = ctrl)
fit_pca

```

```{r}
summary(fit_pca$finalModel)
```
```{r}
varImp(fit_pca)
```

caret centers and scales the data
```{r}
fit_pca$preProcess
```

View the loadings.  These match the $rotation object from prcomp( ).
```{r}

variable_contributions = fit_pca$preProcess$rotation
head(variable_contributions)

```

```{r}

data.frame(variable_contributions) %>%
  gf_point(PC2 ~ PC1)

data.frame(variable_contributions) %>%
  gf_text(PC2 ~ PC1, 
          label = row.names(variable_contributions))

```

```{r}

data.frame(variable_contributions) %>%
  filter(abs(PC1) > .15)

```

Get coordinates of data in PC-space
```{r}

predictor_matrix <- segmentationData2 %>%  
  dplyr::select(-Class) %>% # Remove the response variable
  as.matrix() %>%
  scale() #centers and scales

# The symbol %*% does matrix multiplication
data_in_PC_space = predictor_matrix %*% variable_contributions

# Append the response variable
data_in_PC_space = data.frame(data_in_PC_space, 
                              Class = segmentationData$Class)

```

```{r}

data.frame(data_in_PC_space) %>%
  gf_point(PC2 ~ PC1, col =~ Class)

```

Predicting the first row of the data
```{r}

predict(fit_pca, newdata = segmentationData2[1, ], type = "prob")

```

## Changing the number of PCs
```{r}
set.seed(12)

ctrl = trainControl(method = "cv", number = 5,
                    preProcOptions = list(thresh = 1))

#pcaComp = 5 or thresh = .8

fit_pca_all = train( Class ~ ., 
            data = segmentationData2, 
            method = "glm",
            family = "binomial",
            preProc = c("pca"),
            trControl = ctrl)
fit_pca_all
```

Percentage of variation explained (only if you used thresh = 1 or pcaComp = the maximum)
```{r}

variable_contributions = fit_pca_all$preProcess$rotation
data_in_PC_space = predictor_matrix %*% variable_contributions

var_explained = apply(data_in_PC_space, 2, var)
pve = var_explained/sum(var_explained)
pve

cumsum(pve)

```
Scree plot
```{r}

var_df = data.frame(pve, 
                    PC = 1:length(pve),
                    cum_var = cumsum(pve))

var_df %>%
  gf_point(pve ~ PC) %>%
  gf_line(pve ~ PC)

var_df %>%
  gf_point(cum_var ~ PC) %>%
  gf_line(cum_var ~ PC)

```
### prcomp()

```{r}
seg_predictors = segmentationData2[ , -1] # Remove the response variable

pc_info = prcomp(seg_predictors, scale = TRUE)

biplot(pc_info)
```
A biplot combines the graph of loadings and the graph of data points in PC-space into a single graph.  In a case like this, with many variables, it can be hard to read.  I prefer to make these graphs separately, as shown below.

Graph of variable loadings:
```{r}
PC_loadings = data.frame(pc_info$rotation)

PC_loadings %>%
  gf_point(PC2 ~ PC1)

PC_loadings %>%
  gf_text(PC2 ~ PC1, label =~ rownames(.))
```
Graph of data points in PC-space:
```{r}
data_in_PC_space_from_prcomp = data.frame(pc_info$x, Class = segmentationData$Class)

data_in_PC_space_from_prcomp %>%
  gf_point(PC2 ~ PC1, color =~ Class)

```

### Using a for() loop to do a single layer of 5-fold CV, so we can use prcomp() to do PCA without scaling:
```{r}
num_cells = nrow(segmentationData2)

set.seed(12)
rep_1_5 = rep(1:5, length = num_cells)
cv_folds = sample(rep_1_5, num_cells)

predictions_no_scale = vector(length = num_cells)

for(fold in 1:5){
  train = which(cv_folds != fold)
  test = which(cv_folds == fold)

  # Do PCA on the training set
  pc_info = prcomp(seg_predictors[train, ], center=TRUE, scale=FALSE)

  # Select columns 1-22 for the first 22 PCs, to match fit_pca.
  # Append the response variable.
  prepared_data = data.frame(pc_info$x[ , 1:22], Class = segmentationData2$Class[train])

  # Fit the model to the training set.
  # Could also do this with caret, using no pre-processing and method = "none".
  fit_no_scale = glm(Class ~ ., data = prepared_data[train, ], family = "binomial")

  # Get coordinates of test set in PC-space.
  # Convert to a data frame to enable the next step.
  test_set_pc = data.frame(predict(pc_info, seg_predictors[test, ]))

  # Make predictions using the logistic regression model.
  predictions_no_scale[test] = predict(fit_no_scale, newdata = test_set_pc, type = "response")

} # end 5-fold CV

# Compute accuracy
predictions_class = ifelse(predictions_no_scale > 0.5, "WS", "PS")  # Using a 50% threshold
sum(predictions_class == segmentationData2$Class)/length(segmentationData2$Class)

```
In this case, the accuracy without scaling is similar to the accuracy with scaling.  This is a bit surprising, because there is a wide difference in standard deviations of variables in the original data.

```{r}
std_devs = apply(seg_predictors, 2, sd)
summary(std_devs)
```

### Optional:  PLS
Partial Least Squares is a supervised learning technique that works similarly to using PCA as a preprocessing step before linear regression, but it finds linear combinations of the predictors that best represent the covariance between the predictors and the response (rather than just the variance of the predictors).  This means it can often get good predictions using fewer components than needed with PCA.

If you're interested, check out this video:  https://youtu.be/n_Ll9t6iPvE
```{r}
my_sigma = matrix(c(2,1,1,1), nc = 2)
set.seed(12)
example_1 = data.frame(mvrnorm(200, mu = c(0,0), Sigma = my_sigma))
example_1 <- example_1 %>%
  mutate(y = X2 - X1)

example_1 %>%
  gf_point(X2~X1, col =~ y) 
```
```{r}

data(BloodBrain)
head(logBBB)
head(bbbDescr[,1:6])

dim(bbbDescr)
```

```{r}

bbb_numeric = select_if(bbbDescr, is.numeric)
correlations <-cor(bbb_numeric, use = "pairwise.complete.obs")
corrplot::corrplot(correlations, type="upper", order="hclust",
         col=rev(brewer.pal(n=8, name="RdYlBu")))

```

```{r}

set.seed(12)

ctrl = trainControl(method = "cv", number = 5)
fit_pcr = train(y = logBBB,
                x = bbbDescr, 
                method = "pcr",
                tuneGrid = expand.grid(ncomp = 6:26), #seq(1, 134, by = 5)
                trControl = ctrl)
fit_pcr

```

```{r}

set.seed(12)

ctrl = trainControl(method = "cv", number = 5)
fit_pls = train(y = logBBB,
                x = bbbDescr, 
                method = "pls",
                tuneGrid = expand.grid(ncomp = 6:26),  #seq(1, 134, by = 5)
                trControl = ctrl)
fit_pls

```

```{r}

varImp(fit_pls)

```
```{r}

coef(fit_pls$finalModel)

````

### 2024-12-09 Draft_Cleaning.Rmd
adjusted tuning params

@@
   eta = c(0.1, 0.2, 0.3), 
-  gamma = c(0, 0.1), 
-  colsample_bytree = c(0.6, 0.8), 
-  min_child_weight = c(1, 2, 5), 
+  gamma = 0, 
+  colsample_bytree = 0.8, 
+  min_child_weight =1, 
   subsample = c(0.8, 1)
@@
                    tuneGrid = tune_grid,
-                   verbosity = 0,
                    trControl = ctrl,

### 2024-12-10 Draft_Cleaning.Rmd
continued to test and tune models, split data by river

@@
 library(ggformula)
+library(DiagrammeR)
 
@@
 
+plot(dredge_PCA,  type="l")
+
+autoplot(dredge_PCA)
+
 variable_contributions<- dredge_PCA$rotation
@@
 
-data.frame(variable_contributions) |>
-  gf_point(PC2 ~ PC1)
+# data.frame(variable_contributions) |>
+#   mutate(RIVER = gage_survey_cleaned$RIVER)
+#   gf_point(PC2 ~ PC1)
 
@@
 
+data.frame(variable_contributions) |>
+  filter(abs(PC1) > 0.11)
+
 predictor_matrix <- gage_survey_PCA |>
-  dplyr::select(-VOLUMEDREDGED) %>% # Remove the response variable
-  as.matrix() %>%
+  dplyr::select(-VOLUMEDREDGED) |> # Remove the response variable
+  as.matrix() |>
   scale() #centers and scales
@@
 
-data.frame(data_in_PC_space) %>%
-  gf_point(PC2 ~ PC1)
+data_in_PC_space = data.frame(data_in_PC_space,
+                              River = gage_survey_cleaned$RIVER,
+                              Pool = gage_survey_cleaned$POOL,
+                              DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
+                              Season = gage_survey_cleaned$Season)
+
+data.frame(data_in_PC_space)|>
+  gf_point(PC2 ~ PC1, col = ~ Season)
+
+data.frame(data_in_PC_space)|>
+  gf_point(PC2 ~ PC1, col = ~DREDGE_TYPE)
+
+### Scree Plot
+data_in_PC_space = predictor_matrix %*% variable_contributions
+var_explained = apply(data_in_PC_space,2,var)
+pve = var_explained/sum(var_explained)
+pve
+
+cumsum(pve)
+
+var_df <- data.frame(pve, PC = 1:length(pve),
+                     cum_var = cumsum(pve))
+
+var_df|>
+  gf_point(cum_var ~ PC) |>
+  gf_line(cum_var ~ PC)
+
+pca_scores <-dredge_PCA$x
+pca_loadings <- dredge_PCA$rotation
 
@@
 library(xgboost)
-gage_xgb <- gage_survey |>
+gage_xgb <- gage_survey_cleaned |>
   select(-lead_7, -lead_14,-DATE_START,
@@
                  method = "xgbTree",
-                 tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
-                                        eta = 0.3, gamma = 0,
-                                        colsample_bytree = .8,
+               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
+                                              max_depth = 1:10,
+                                        eta =c(0.3), gamma = 0,
+                                        colsample_bytree = 1,
                                         min_child_weight = 1, subsample = 1),
@@
 varImp(fit_dredge)
+xgb.plot.tree(model = fit_dredge$finalModel, trees = 0)
 
@@
                        method = "xgbTree",
-                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
-                                        eta = 0.3, gamma = 0,
+                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
+                                              max_depth = 1:10,
+                                        eta =c(0.3), gamma = 0,
                                         colsample_bytree = 1,
@@
 }
-
-tune_grid <- expand.grid(
-  nrounds = c(50, 100, 150), 
-  max_depth = c(3, 6, 9), 
-  eta = c(0.1, 0.2, 0.3), 
-  gamma = 0, 
-  colsample_bytree = 0.8, 
-  min_child_weight =1, 
-  subsample = c(0.8, 1)
-)
-
-set.seed(123)
-ctrl = trainControl(method = "cv", number = 10, savePredictions = "final")
-
-fit_dredge2 = train(VOLUMEDREDGED ~ .,
-                   data = gage_xgb,
-                   method = "xgbTree",
-                   tuneGrid = tune_grid,
-                   trControl = ctrl,
-                   na.action = na.exclude)
-
-# Model performance
-fit_dredge2
-
-# Model evaluation
-predictions <- fit_dredge2$pred  # Predictions from cross-validation
-head(predictions)
-
-# RMSE, R-Squared, etc.
-cv_rmse <- sqrt(mean((predictions$obs - predictions$pred)^2))
-cv_r2 <- cor(predictions$obs, predictions$pred)^2
-
-cat("RMSE: ", cv_rmse, "\n")
-cat("R-Squared: ", cv_r2, "\n")
 
@@
                                          # assessment of accuracy
+```
 
-dredge_data_cleaning|>
-  group_by(POOL)|>
-  summarize(count())
+## Well that didnt work! Lets try modeling between the two river systems 
+```{r}
+combined_data <- bind_rows(survey_Data, dredge_data_cleaning)
+
+combined_data <-combined_data |>
+     filter(DATE_START >= as.Date("1999-04-01") &
+         DATE_START <= as.Date("2021-09-30"))
+
+combined_data <-combined_data |>
+  mutate(lead_7 = (DATE_START - 7),
+         lead_14 = (DATE_START - 14))
+Nas<- combined_data|>
+  filter(if_any(everything(),is.na))
+
+gage_survey <- combined_data |>
+  left_join(gage_data2, by = "DATE_START")
+
+Illinois_River <- gage_survey |>
+  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
+         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
+         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
+         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
+  filter(RIVER == "Illinois_Waterway")
+
+Miss_River <- gage_survey |>
+  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
+         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
+         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
+         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
+         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
+  filter(RIVER == "Mississippi_River")
+
+Miss_full <- Miss_River|>  
+  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+
+IWW_full <- Illinois_River|>  
+  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+
+Miss_Cleaned <- Miss_full |>
+  filter(complete.cases(Miss_full))
+
+IWW_Cleaned <- IWW_full |>
+  filter(complete.cases(IWW_full))
+
+IWW_Cleaned <- IWW_Cleaned |>
+  mutate("Season" = sapply(DATE_START,date_to_season))
+
+Miss_Cleaned <- Miss_Cleaned |>
+  mutate("Season" = sapply(DATE_START,date_to_season))
+
 ```
+
+## xgboost split by river
+#Miss
+```{r}
+Miss_xgb <-Miss_Cleaned |>
+  select(-lead_7, -lead_14,-DATE_START,
+         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
+         -DREDGINGPURPOSE, -RIVER)|>
+    mutate_if(is.character, factor)
+
+ctrl = trainControl(method = "cv", number = 10) 
+Miss_xgb_fit = train(VOLUMEDREDGED ~ .,
+                 data = Miss_xgb,
+                 method = "xgbTree",
+               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
+                                              max_depth = 1:10,
+                                        eta =c(0.3), gamma = 0,
+                                        colsample_bytree = 1,
+                                        min_child_weight = 1, subsample = 1),
+                 verbosity = 0,
+                 trControl = ctrl,
+                 na.action = na.exclude)
+
+varImp(Miss_xgb_fit)
+xgb.plot.tree(model = Miss_xgb_fit$finalModel, trees = 0)
+
+Miss_xgb_fit
+
+import_matrix = xgb.importance(model = Miss_xgb_fit$finalModel)
+import_matrix
+xgb.plot.importance(import_matrix)
+```
+
+#IWW
+```{r}
+IWW_xgb <-IWW_Cleaned |>
+  select(-lead_7, -lead_14,-DATE_START,
+         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
+         -DREDGINGPURPOSE, -RIVER)|>
+    mutate_if(is.character, factor)
+
+ctrl = trainControl(method = "cv", number = 10) 
+IWW_xgb_fit = train(VOLUMEDREDGED ~ .,
+                 data = IWW_xgb,
+                 method = "xgbTree",
+               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
+                                              max_depth = 1:10,
+                                        eta =c(0.3), gamma = 0,
+                                        colsample_bytree = 1,
+                                        min_child_weight = 1, subsample = 1),
+                 verbosity = 0,
+                 trControl = ctrl,
+                 na.action = na.exclude)
+
+varImp(IWW_xgb_fit)
+xgb.plot.tree(model = IWW_xgb_fit$finalModel, trees = 0)
+
+IWW_xgb_fit
+
+import_matrix = xgb.importance(model = IWW_xgb_fit$finalModel)
+import_matrix
+xgb.plot.importance(import_matrix)
+```

### 2024-12-10 Draft_Cleaning.Rmd
split by rivers for PCA, clean PCA plots, got rid of unneeded doc

@@
 library(DiagrammeR)
+```
 
-dredge_data<-read_csv("DredgeData.csv")
-survey_data<-read_csv("SurveyData.csv")
-gage_data<-read_csv("UMR_IWW_1999_2021.csv")
+Read in the data
+```{r}
+dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
+survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
+gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)
 
-gage_data2<- gage_data |>
+### The gage data has multiple records across different gages per day, we need to
+### combine the rows by date so that there's only one record for date across all gages.
+### Since we're going to be doing a lot of work with the date column, lets format it as 
+### a date object and arrange by DATE_START and remove the old Date column
+gage_data<- gage_data |>
   group_by(Date)|>
@@
   mutate(DATE_START= as_date(dmy(Date)))|>
-    arrange(DATE_START)
+    arrange(DATE_START)|>
+  select(-Date)
 
-gage_data2<-gage_data2|>
-    mutate(across(c(-Date,-DATE_START),
+### Some of the gage data has observation gaps where the sensors may not have been
+### working, lets interpolate gage values where there's data before and after a gap
+### with a maximum window of 2
+gage_data<-gage_data|>
+    mutate(across(c(-DATE_START),
                 ~ ifelse(is.na(.),
                          coalesce((lag(.) + lead(.)) / 2, lag(.), lead(.)),
-                         .)))|>
-  select(-Date)
+                         .)))
+```
 
-survey_Data<- survey_data |>
+Now lets clean the survey data. I only want conditional survey records between
+1999 to 2000, these are going to act as my background data. I want to combine
+these records with my dredge data. I'm also going to filter by river, remove
+pools that are not of interest, clean up the SURVEYJOBIDPK, and create some 
+columns to support my join. Dates are also formatted as date objects and some
+columns are renamed to make it easier to join my surveys to my dredge data.
+Finally, select the columns we need for our analysis to join to our dredging 
+data. 
+```{r}
+survey_data<- survey_data |>
   filter(SURVEYTYPE == "conditional",
@@
           RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
-         SURVEYJOBIDPK == str_replace_all(SURVEYJOBIDPK, "IL", "IW"),
          !POOL %in% c("LP","CS"))|>
@@
          )|>
-  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,UP_RIV_MIL,DN_RIV_MIL)
+  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
+         UP_RIV_MIL,DN_RIV_MIL)
+```
 
-dredge_data_cleaning <- dredge_data |>
+Lets clean the dredge data, we only want records that are channel dredging events
+and since the dredge records only go back to 1999, we just need to filter out data beyond 2022.
+```{r}
+dredge_data <- dredge_data |>
    filter(
@@
   mutate(
-    river_code = str_sub(PREHYDROSURVEY_ID, 1, 2),
-    pool_code = str_sub(PREHYDROSURVEY_ID, 4, 5),
-    district_code = str_sub(PREHYDROSURVEY_ID, 7, 9),
-    river_mile = str_sub(PREHYDROSURVEY_ID, 23, 30),
-    year_iteration = str_sub(PREHYDROSURVEY_ID, 32,34),
-    SURVEYJOBIDPK = ifelse(
-      str_detect(PREHYDROSURVEY_ID, "MVRUM|MVRIW"),
-      PREHYDROSURVEY_ID,
-      paste0(district_code, river_code, pool_code, "_", river_mile, "_", year_iteration)  
-    ),
-      SURVEYJOBIDPK = str_replace_all(SURVEYJOBIDPK, "IL", "IW"),
-     DATE_START = as_date(mdy_hm(DATE_START)),
-    SDSFEATURENAME = FEATURENAME)|>
-  select(SURVEYJOBIDPK,EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,DATE_START, UP_RIV_MIL,DN_RIV_MIL)
+     DATE_START = as_date(mdy_hm(DATE_START)))|>
+  select(EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
+         DATE_START, UP_RIV_MIL,DN_RIV_MIL)
+```
 
-combined_data <- bind_rows(survey_Data, dredge_data_cleaning)
+Combine the dredge data and survey data into 1 table. The survey data will act as background/non-dredge data.
+```{r}
+combined_data <- bind_rows(survey_data, dredge_data)
 
@@
 gage_survey <- combined_data |>
-  left_join(gage_data2, by = "DATE_START")
+  left_join(gage_data, by = "DATE_START")
 
@@
 gage_survey_full <- gage_survey_filt|>  
-  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
-  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
 
@@
 
-### Create factored season variable that could be used in analysis
+```
+
+Create factored season variable that could be used in analysis
+```{r}
 date_to_season <- function(date) {
@@
 
+dredge <-gage_survey_cleaned |>
+  filter(DREDGINGPURPOSE != "None")
+  hist(dredge$VOLUMEDREDGED)
+
+  median(dredge$VOLUMEDREDGED)
+
+hist(gage_survey_cleaned$VOLUMEDREDGED)
+
 plot(dredge_PCA,  type="l")
 
-autoplot(dredge_PCA)
+autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'SEASON',loadings = TRUE,
+         loadings.colour = 'black',
+         loadings.label = TRUE, loadings.label.size = 4, 
+         loadings.label.colour = "blue")
 
@@
 
-ctrl = trainControl(method = "cv", number = 10) 
+ctrl = trainControl(method = "cv", number = 5) 
 fit_dredge = train(VOLUMEDREDGED ~ .,
@@
 n = nrow(gage_xgb)
-n_outer_folds = 10
+n_outer_folds = 5
 
@@
 
-  ctrl = trainControl(method = "cv", number = 10)
+  ctrl = trainControl(method = "cv", number = 5)
   fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
@@
 ## Well that didnt work! Lets try modeling between the two river systems
+
+### Cleaning the data 
 ```{r}
@@
 Miss_full <- Miss_River|>  
-  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
-  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
 
 IWW_full <- Illinois_River|>  
-  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
-  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
 
@@
   mutate("Season" = sapply(DATE_START,date_to_season))
+
+```
+
+## PCA
+```{r}
 

### 2024-12-10 Draft_Cleaning.Rmd
cleaning up code, making final figures

@@
 library(DiagrammeR)
+library(gridExtra)
+
+### Key Functions 
+# Function to plot PCA
+plot_pca <- function(PCAvalues, top_loadings, PC1_explained, PC2_explained, grouping_var) {
+
+  p <- ggplot(PCAvalues, aes(x = "PC1", y = "PC2", colour = grouping_var)) +
+    geom_point(size = 2) +  # Scatter plot
+    geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+                 arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
+    annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+             label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
+    theme_minimal() +
+    labs(title = paste("PCA Plot with Top 10 Loadings for PC1 and PC2 - Grouped by", grouping_var),
+         x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+         y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+  return(p)
+}
+
+#Function to calculate season from date 
+date_to_season <- function(date) {
+  month <- as.numeric(format(date, "%m"))
+  day <- as.numeric(format(date, "%d"))
+  # Define the seasons based on date ranges
+  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
+    return("Spring")
+  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
+    return("Summer")
+  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
+    return("Fall")
+  } else {
+    return("Winter")
+  }
+}
+
 ```
 
-Read in the data
+### Read in the data
 ```{r}
@@
 gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)
+```
 
-### The gage data has multiple records across different gages per day, we need to
-### combine the rows by date so that there's only one record for date across all gages.
-### Since we're going to be doing a lot of work with the date column, lets format it as 
-### a date object and arrange by DATE_START and remove the old Date column
+### Clean the gage data 
+The gage data has multiple records across different gages per day, we need to
+combine the rows by date so that there's only one record for date across all gages.
+Since we're going to be doing a lot of work with the date column, lets format it as 
+a date object and arrange by DATE_START and remove the old Date column
+```{r}
 gage_data<- gage_data |>
@@
 
+### Cleaning the data
 Now lets clean the survey data. I only want conditional survey records between
@@
 these records with my dredge data. I'm also going to filter by river, remove
-pools that are not of interest, clean up the SURVEYJOBIDPK, and create some 
+pools that are not of interest, and create some 
 columns to support my join. Dates are also formatted as date objects and some
@@
      filter(DATE_START >= as.Date("1999-04-01") &
-         DATE_START <= as.Date("2021-09-30"))
+         DATE_START <= as.Date("2021-09-30")) |>
+          mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
+                                      DREDGINGPURPOSE == "channel" ~ "Dredged"))
 
@@
          lead_14 = (DATE_START - 14))
+```
+
+Check for NAs for PCA analysis. 
+```{r}
 Nas<- combined_data|>
@@
   filter(complete.cases(gage_survey))
+```
 
+Join the gage data by the 7 day prior observations and by the 14 day prior observations
+```{r}
 gage_survey_full <- gage_survey_filt|>  
@@
   left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+```
 
+Check again for NAs, remove any remaining NA rows. 
+```{r}
 Nas<- gage_survey_full|>
@@
 ```{r}
-date_to_season <- function(date) {
-  month <- as.numeric(format(date, "%m"))
-  day <- as.numeric(format(date, "%d"))
-  # Define the seasons based on date ranges
-  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
-    return("Spring")
-  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
-    return("Summer")
-  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
-    return("Fall")
-  } else {
-    return("Winter")
-  }
-}
-
 gage_survey_cleaned <- gage_survey_cleaned |>
@@
 ```
+Now on to analysis!
 
+To first explore the variables, we will use PCA.
+### PCA
+PCA using prcomp
 ```{r}
@@
 select(where(is.numeric), -EXECYEAR)
-
-dredge_numeric<-gage_survey_cleaned|>
-   select(where(is.numeric), -EXECYEAR)
-correlations <-cor(dredge_numeric, use = "pairwise.complete.obs")
-corrplot::corrplot(correlations)
 
@@
 
-biplot(dredge_PCA)
-
-dredge <-gage_survey_cleaned |>
-  filter(DREDGINGPURPOSE != "None")
-  hist(dredge$VOLUMEDREDGED)
-
-  median(dredge$VOLUMEDREDGED)
-
-hist(gage_survey_cleaned$VOLUMEDREDGED)
-
 plot(dredge_PCA,  type="l")
 
-autoplot(dredge_PCA,data = gage_survey_cleaned, 
-         color = 'SEASON',loadings = TRUE,
-         loadings.colour = 'black',
-         loadings.label = TRUE, loadings.label.size = 4, 
-         loadings.label.colour = "blue")
+a<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'Season',loadings = FALSE)+
+  ggtitle("Season")
 
+b<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'POOL',loadings= FALSE)+
+  ggtitle("River Pool")+
+    guides(color = guide_legend(ncol = 2))
+
+c<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'RIVER',loadings = FALSE)+
+  ggtitle("River")
+
+d<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'DREDGINGPURPOSE',loadings = FALSE)+
+        ggtitle("Dredging Purpose")+
+  labs(color = "Dredging?")
+
+ggsave("Initial_Full_PCA.png",grid.arrange(a,b,c,d,ncol = 2),width = 10, height = 8)
 variable_contributions<- dredge_PCA$rotation
 head(variable_contributions)
+```
 
+PCA using Caret
+```{r}
 set.seed(12)
@@
 head(variable_contributions)
-
-# data.frame(variable_contributions) |>
-#   mutate(RIVER = gage_survey_cleaned$RIVER)
-#   gf_point(PC2 ~ PC1)
 
@@
                               DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
-                              Season = gage_survey_cleaned$Season)
+                              Season = gage_survey_cleaned$Season,
+                              Dredge_Volume = gage_survey_cleaned$VOLUMEDREDGED)
 
@@
   gf_line(cum_var ~ PC)
-
-pca_scores <-dredge_PCA$x
-pca_loadings <- dredge_PCA$rotation
-
 ```
@@
   select(-lead_7, -lead_14,-DATE_START,
-         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
+         -EXECYEAR,-DREDGE_TYPE,
          -DREDGINGPURPOSE)|>
@@
 xgb.plot.importance(import_matrix)
+
 ```
@@
 ```{r}
-combined_data <- bind_rows(survey_Data, dredge_data_cleaning)
+combined_data <- bind_rows(survey_data, dredge_data)
 
@@
 
+gage_IWW <- gage_data |>
+  select(DATE_START,BEAI2,HAVI2,
+         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
+         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2)
+
+gage_Miss <- gage_data |>
+  select(DATE_START,BRLI4,CMMI4, DBQI4,
+         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
+         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
+         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2)
+
 Illinois_River <- gage_survey |>
-  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
+  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
          POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
@@
 Miss_River <- gage_survey |>
-  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
+  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
          POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
@@
 Miss_full <- Miss_River|>  
-  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
-  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+  left_join(gage_Miss, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_Miss, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
 
 IWW_full <- Illinois_River|>  
-  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
-  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
+  left_join(gage_IWW, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
+  left_join(gage_IWW, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
 
@@
 
-## PCA
+## IWW PCA
 ```{r}
+IWW_PCA <- IWW_Cleaned |>
+select(where(is.numeric), -EXECYEAR)
+
+IWW_prcomp <- IWW_PCA|>
+  select(-VOLUMEDREDGED)
+
+IWW_PCA = prcomp(IWW_prcomp, scale = T)
+
+biplot(IWW_PCA)
+
+plot(IWW_PCA,  type="l")
+
+autoplot(IWW_PCA,data = IWW_Cleaned, 
+         color = 'Season',loadings = TRUE,
+         loadings.colour = 'black',
+         loadings.label = TRUE, loadings.label.size = 4, 
+         loadings.label.colour = "blue")
+
+PCAvalues <- data.frame(Pool = IWW_Cleaned$POOL,
+                        Dredge_Type = IWW_Cleaned$DREDGE_TYPE,
+                        Season = IWW_Cleaned$Season,
+                        Dredge_Occurance = IWW_Cleaned$DREDGINGPURPOSE,
+                        IWW_PCA$x)
+
+explained_variance <- (IWW_PCA$sdev^2) / sum(IWW_PCA$sdev^2) * 100
+
+# Extract PC1 and PC2 explained variance
+PC1_explained <- explained_variance[1]
+PC2_explained <- explained_variance[2]
+
+# Extract loadings of the variables
+PCAloadings <- data.frame(Variables = rownames(IWW_PCA$rotation), IWW_PCA$rotation)
+
+PCAloadings_PC1 <- PCAloadings|>
+  arrange(desc(abs(PC1))) |>
+  slice_head(n=10)
+
+PCAloadings_PC2 <- PCAloadings|>
+  arrange(desc(abs(PC2))) |>
+  slice_head(n=10)
+
+top_loadings <- bind_rows(PCAloadings_PC1, PCAloadings_PC2)
+
+# Plot PCA with top 10 loadings
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Season)) +
+  geom_point(size = 2) +
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Pool)) +
+  geom_point(size = 2) +
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Type)) +
+  geom_point(size = 2) + 
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") + 
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Occurance)) +
+  geom_point(size = 2) + 
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) + 
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+```
+
+## Miss PCA 
+```{r}
+Miss_PCA <- Miss_Cleaned |>
+select(where(is.numeric), -EXECYEAR)
+
+Miss_prcomp <- Miss_PCA|>
+  select(-VOLUMEDREDGED)
+
+Miss_PCA = prcomp(Miss_prcomp, scale = T)
+
+biplot(Miss_PCA)
+biplot(Miss_PCA, showLoadings = TRUE,)
+
+plot(Miss_PCA,  type="l")
+
+PCAvalues <- data.frame(Pool = Miss_Cleaned$POOL,
+                        Dredge_Type = Miss_Cleaned$DREDGE_TYPE,
+                        Season = Miss_Cleaned$Season,
+                        Dredge_Occurance = Miss_Cleaned$DREDGINGPURPOSE,
+                        Miss_PCA$x)
+
+explained_variance <- (Miss_PCA$sdev^2) / sum(Miss_PCA$sdev^2) * 100
+
+# Extract PC1 and PC2 explained variance
+PC1_explained <- explained_variance[1]
+PC2_explained <- explained_variance[2]
+
+# Extract loadings of the variables
+PCAloadings <- data.frame(Variables = rownames(Miss_PCA$rotation), Miss_PCA$rotation)
+
+PCAloadings_PC1 <- PCAloadings|>
+  arrange(desc(abs(PC1))) |>
+  slice_head(n=10)
+
+PCAloadings_PC2 <- PCAloadings|>
+  arrange(desc(abs(PC2))) |>
+  slice_head(n=10)
+
+top_loadings <- bind_rows(PCAloadings_PC1, PCAloadings_PC2)
+
+# Plot PCA with top 10 loadings
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Season)) +
+  geom_point(size = 2) +  # Scatter plot
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Pool)) +
+  geom_point(size = 2) +  # Scatter plot
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
+
+ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Type)) +
+  geom_point(size = 2) +  # Scatter plot
+  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
+               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
+  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
+           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
+  theme_minimal() +
+  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
+      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
+       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
 
@@
   select(-lead_7, -lead_14,-DATE_START,
-         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
+         -EXECYEAR,-DREDGE_TYPE,
          -DREDGINGPURPOSE, -RIVER)|>

### 2025-04-29 Draft_Cleaning.Rmd
reran some draft cleaning stuff

@@
 library(gridExtra)
+library(keras)
+library(tensorflow)
 
@@
 survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
-gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)
+gage_data<-read_csv("UMR_IWW_1999_2024.csv", show_col_types = FALSE)
 ```
@@
 Now lets clean the survey data. I only want conditional survey records between
-1999 to 2000, these are going to act as my background data. I want to combine
+1999 to 2022, these are going to act as my background data. I want to combine
 these records with my dredge data. I'm also going to filter by river, remove
@@
   filter(SURVEYTYPE == "conditional",
-         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
+         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2024,
           RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
@@
           DREDGINGPURPOSE != "harbor",
-           EXECYEAR <= 2022)|>
+           EXECYEAR <= 2024)|>
   mutate(
@@
      filter(DATE_START >= as.Date("1999-04-01") &
-         DATE_START <= as.Date("2021-09-30")) |>
+         DATE_START <= as.Date("2024-12-31")) |>
           mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
@@
   filter(complete.cases(gage_survey))
+
+write_csv(gage_survey_filt,"Gage_Survey_Dredge.csv")
 ```
@@
 ```
+
+Lets make some plots of the data!
+```{r}
+gages_pivoted <- gage_data|>
+    pivot_longer(cols = -DATE_START, 
+               names_to = "gage", 
+               values_to = "value")
+
+gage_survey_cleaned |>
+  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START))+
+  geom_line()
+
+gages_pivoted |>
+  ggplot(aes(y = value, x = DATE_START, color = gage, group = gage)) +
+  geom_line(size = 1) +  # Adjust line thickness
+  scale_color_viridis_d() +  # A color scale that's visually appealing
+  labs(
+    title = "Time Series of Gage Values",
+    x = "Date",
+    y = "Value",
+    color = "Gage"
+  ) +  # Adding titles and labels
+  theme_minimal() +  # Use a minimal theme
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
+    axis.title = element_text(size = 12),  # Increase title font size
+    legend.position = "right",  # Position the legend on the right
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 10)
+  )
+```
+
+```{r}
+gage_survey |>
+  filter(RIVER == "Mississippi_River")|>
+  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START,color = POOL))+
+  geom_point()+
+  labs(title = "Miss Dredging")
+
+gage_survey |>
+  filter(RIVER == "Illinois_Waterway")|>
+  ggplot(aes(y=VOLUMEDREDGED, x=DATE_START, color = POOL))+
+  geom_point()+
+  labs(title = "IWW Dredging")
+
+```
+
 Now on to analysis!
@@
 
-fit_pca = train(VOLUMEDREDGED ~ .,
+fit_pca = caret::train(VOLUMEDREDGED ~ .,
             data = gage_survey_PCA,
@@
 ctrl = trainControl(method = "cv", number = 5) 
-fit_dredge = train(VOLUMEDREDGED ~ .,
+fit_dredge = caret::train(VOLUMEDREDGED ~ .,
                  data = gage_xgb,
@@
   ctrl = trainControl(method = "cv", number = 5)
-  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
+  fit_dredge_DCV = caret::train(VOLUMEDREDGED ~ .,
                        data = gage_xgb[in_train, ],
@@
      filter(DATE_START >= as.Date("1999-04-01") &
-         DATE_START <= as.Date("2021-09-30"))
+         DATE_START <= as.Date("2024-12-31"))
 
@@
          HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
-         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2)
+         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2,"05552500","05527500",
+         "05585000","05583000","05570000","05555300")
 
@@
          MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
-         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2)
+         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2,"05446500")
 
@@
          HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
-         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
+         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2,"05552500","05527500",
+         "05585000","05583000","05570000","05555300") |>
   filter(RIVER == "Illinois_Waterway")
@@
          MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
-         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
+         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2,"05446500") |>
   filter(RIVER == "Mississippi_River")
@@
 ctrl = trainControl(method = "cv", number = 10) 
-Miss_xgb_fit = train(VOLUMEDREDGED ~ .,
+Miss_xgb_fit = caret::train(VOLUMEDREDGED ~ .,
                  data = Miss_xgb,
@@
 ctrl = trainControl(method = "cv", number = 10) 
-IWW_xgb_fit = train(VOLUMEDREDGED ~ .,
+IWW_xgb_fit = caret::train(VOLUMEDREDGED ~ .,
                  data = IWW_xgb,
@@
 ```
+###LSTM
+```{r}
+library(tensorflow)
+library(keras)
+model_7 <- keras_model_sequential() |>
+  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(7, ncol(gage_xgb) - 1))  |>
+  layer_dropout(rate = 0.2)|>
+  layer_lstm(units = 50, return_sequences = TRUE) |>
+  layer_dropout(rate = 0.2) |>
+  layer_lstm(units = 50) |>
+  layer_dropout(rate = 0.2) |>
+  layer_dense(units = 1)
+
+model_7 %>% compile(
+  loss = 'mean_squared_error',
+  optimizer = 'adam'
+)
+
+# Function to create sequences based on window size
+create_sequences <- function(data, target_column, sequence_length) {
+  X <- list()
+  y <- list()
+
+  for (i in (sequence_length + 1):nrow(data)) {
+    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])  # Features excluding target
+    y[[i - sequence_length]] <- data[i, target_column]  # Target column
+  }
+
+  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))  # Reshape to 3D array
+  y <- unlist(y)
+
+  return(list(X = X, y = y))
+}
+
+# Example: Create sequences for 7-day and 14-day windows
+window_size_7 <- 7
+window_size_14 <- 14
+
+# Load necessary libraries
+library(recipes)
+
+# Assume gage_xgb is your original dataset
+gage_xgb_scaled_recipe <- recipe(~ ., data = gage_xgb) %>%
+  # Define step to scale all numeric predictors
+  step_scale(all_numeric_predictors()) %>%
+  # You can also center them if needed
+  step_center(all_numeric_predictors())
+
+# Prep the recipe
+gage_xgb_prepped <- prep(gage_xgb_scaled_recipe, training = gage_xgb)
+
+# Apply the scaling to the data
+gage_xgb_scaled <- bake(gage_xgb_prepped, new_data = gage_xgb)
+
+# View scaled data
+head(gage_xgb_scaled)
+
+# Assume `gage_xgb_scaled` is your scaled dataset and the target column is 'VOLUMEDREDGED'
+data_windows_7 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_7)
+data_windows_14 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_14)
+
+# X_train and y_train for 7-day window
+X_train_7 <- data_windows_7$X
+y_train_7 <- data_windows_7$y
+
+# X_train and y_train for 14-day window
+X_train_14 <- data_windows_14$X
+y_train_14 <- data_windows_14$y
+
+# Define the train-test split
+train_size_7 <- floor(0.8 * length(y_train_7))  # 80% for training
+train_size_14 <- floor(0.8 * length(y_train_14))  # 80% for training
+
+# Training and Testing Data for 7-day window
+X_train_7_final <- X_train_7[1:train_size_7,,]
+y_train_7_final <- y_train_7[1:train_size_7]
+X_test_7_final <- X_train_7[(train_size_7+1):length(y_train_7),,]
+y_test_7_final <- y_train_7[(train_size_7+1):length(y_train_7)]
+
+# Training and Testing Data for 14-day window
+X_train_14_final <- X_train_14[1:train_size_14,,]
+y_train_14_final <- y_train_14[1:train_size_14]
+X_test_14_final <- X_train_14[(train_size_14+1):length(y_train_14),,]
+y_test_14_final <- y_train_14[(train_size_14+1):length(y_train_14)]
+
+# Train the model for the 7-day window
+history_7 <- model_7 %>% fit(
+  X_train_7_final, 
+  y_train_7_final,
+  epochs = 20, 
+  batch_size = 32, 
+  validation_data = list(X_test_7_final, y_test_7_final)
+)
+
+# Evaluate model on 7-day window
+score_7 <- model_7 %>% evaluate(X_test_7_final, y_test_7_final)
+cat('Test loss for 7-day window:', score_7, "\n")
+
+predictions_7 <- model_7 %>% predict(X_test_7_final)
+```

### 2025-04-29 CSAT_Wrangling.R
updated csat_wrangling script

@@ -1,0 +1,230 @@
+library(readr)
+library(dplyr)
+library(stringr)
+library(purrr)
+library(ggplot2)
+library(lubridate)
+library(tidyverse)
+library(zoo)
+options(scipen = 100, digits = 2)
+
+# Set folder path
+folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101__to__20241231"
+
+reach_data<-read_csv("C:/workspace/DRAFT/ReachData.csv")
+
+# List matching files (ends with dredged_date_vol.csv)
+file_list <- list.files(path = folder_path,
+                        pattern = "_dredged_date_vol\\.csv$",
+                        full.names = TRUE)
+
+# Function to extract metadata and read file
+read_and_annotate <- function(file_path) {
+  # Extract file name only
+  file_name <- basename(file_path)
+  
+  # Use regex to extract River, Pool, and Reach
+  # Pattern: CEMVR_(River)_(Pool)_MVR_(Reach)_dredged_date_vol
+  matches <- str_match(file_name, "^CEMVR_([A-Z]{2})_([A-Z0-9]+)_MVR_([0-9]+)_dredged_date_vol\\.csv$")
+  
+  if (is.na(matches[1,1])) {
+    warning(paste("Filename does not match expected pattern:", file_name))
+    return(NULL)
+  }
+
+  
+  reachname <- 
+  river <- matches[1,2]
+  pool <- matches[1,3]
+  reach <- matches[1,4]
+  reachname <-paste0("CEMVR_",river,"_",pool,"_MVR_",reach) 
+  
+  # Read data
+  df <- read_csv(file_path, show_col_types = FALSE)
+  
+  # Add extracted info
+  df <- df %>%
+    mutate(channelreachidpk = reachname,
+           River = river,
+           Pool = pool,
+           Reach = reach)
+  
+  return(df)
+}
+
+# Apply to all files and combine
+combined_data <- file_list %>%
+  lapply(read_and_annotate) %>%
+  bind_rows()
+
+combined_data<-rename(combined_data, VolumeChange = `Volume Change (cy)`)
+
+data_cleaned <- combined_data|>
+  filter(VolumeChange>500)
+
+
+CSAT_dredge_data <- data_cleaned|>
+  left_join(reach_data, join_by(channelreachidpk))|>
[…]

### 2025-06-26 CSAT_Wrangling.R
removed unneeded files and docs

@@ -1,230 +1,0 @@
-library(readr)
-library(dplyr)
-library(stringr)
-library(purrr)
-library(ggplot2)
-library(lubridate)
-library(tidyverse)
-library(zoo)
-options(scipen = 100, digits = 2)
-
-# Set folder path
-folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101__to__20241231"
-
-reach_data<-read_csv("C:/workspace/DRAFT/ReachData.csv")
-
-# List matching files (ends with dredged_date_vol.csv)
-file_list <- list.files(path = folder_path,
-                        pattern = "_dredged_date_vol\\.csv$",
-                        full.names = TRUE)
-
-# Function to extract metadata and read file
-read_and_annotate <- function(file_path) {
-  # Extract file name only
-  file_name <- basename(file_path)
-  
-  # Use regex to extract River, Pool, and Reach
-  # Pattern: CEMVR_(River)_(Pool)_MVR_(Reach)_dredged_date_vol
-  matches <- str_match(file_name, "^CEMVR_([A-Z]{2})_([A-Z0-9]+)_MVR_([0-9]+)_dredged_date_vol\\.csv$")
-  
-  if (is.na(matches[1,1])) {
-    warning(paste("Filename does not match expected pattern:", file_name))
-    return(NULL)
-  }
-
-  
-  reachname <- 
-  river <- matches[1,2]
-  pool <- matches[1,3]
-  reach <- matches[1,4]
-  reachname <-paste0("CEMVR_",river,"_",pool,"_MVR_",reach) 
-  
-  # Read data
-  df <- read_csv(file_path, show_col_types = FALSE)
-  
-  # Add extracted info
-  df <- df %>%
-    mutate(channelreachidpk = reachname,
-           River = river,
-           Pool = pool,
-           Reach = reach)
-  
-  return(df)
-}
-
-# Apply to all files and combine
-combined_data <- file_list %>%
-  lapply(read_and_annotate) %>%
-  bind_rows()
-
-combined_data<-rename(combined_data, VolumeChange = `Volume Change (cy)`)
-
-data_cleaned <- combined_data|>
-  filter(VolumeChange>500)
-
-
-CSAT_dredge_data <- data_cleaned|>
-  left_join(reach_data, join_by(channelreachidpk))|>
[…]

### 2025-06-26 Draft_Cleaning.Rmd
testing different LSTM configs

@@
 ```{r}
+CSAT_dredge = read_csv("C:/workspace/DRAFT/CSAT_Data/Dredged_Date_Vol.csv",show_col_types = FALSE)
 dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
@@
   mutate(VOLUMEDREDGED = 0,
-         DREDGINGPURPOSE = "None",
-         DREDGE_TYPE = "None",
-         LABOR_TYPE = LABOR_SOURCE,
          EXECYEAR = SURVEY_YEAR,
-         FEATURENAME = SDSFEATURENAME,
          DATE_START = as.character(DATE_START),
@@
          )|>
-  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
+  select(EXECYEAR,DATE_START,VOLUMEDREDGED, RIVER,POOL, 
+         UP_RIV_MIL,DN_RIV_MIL)
+```
+
+```{r}
+CSAT_dredge_clean <- CSAT_dredge |>
+  mutate(VOLUMEDREDGED = VolumeChange,
+         EXECYEAR = year(Date),
+         DATE_START = SurveyDate,
+         RIVER = case_when(River == "UM" ~"Mississippi_River",
+                           River =="IL" ~ "Illinois_Waterway"), 
+         POOL = Pool,
+         UP_RIV_MIL = mileagestart,
+         DN_RIV_MIL = mileageend)|>
+  select(EXECYEAR,DATE_START,VOLUMEDREDGED,RIVER,POOL, 
          UP_RIV_MIL,DN_RIV_MIL)
@@
 ```{r}
-combined_data <- bind_rows(survey_data, dredge_data)
+combined_data <- bind_rows(survey_data, CSAT_dredge_clean)
 
@@
          DATE_START <= as.Date("2024-12-31")) |>
-          mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
-                                      DREDGINGPURPOSE == "channel" ~ "Dredged"))
+           mutate( 
+          #DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
+          #                             DREDGINGPURPOSE == "channel" ~ "Dredged"),
+                  week_of_year = week(DATE_START))
 
@@
 
+gage_survey_full = gage_survey_filt
 write_csv(gage_survey_filt,"Gage_Survey_Dredge.csv")
@@
 ```{r}
+gage_survey_filt <- gage_survey_filt |>
+  mutate(
+    lead_7 = as.Date(lead_7),
+    lead_14 = as.Date(lead_14)
+  )
+
+gage_data <- gage_data |>
+  mutate(DATE_START = as.Date(DATE_START))
+
 gage_survey_full <- gage_survey_filt|>  
@@
   )
+
+gages_pivoted |>
+  ggplot(aes())
 ```
@@
 
-d<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
-         color = 'DREDGINGPURPOSE',loadings = FALSE)+
-        ggtitle("Dredging Purpose")+
+# d<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+#          color = 'DREDGINGPURPOSE',loadings = FALSE)+
+#         ggtitle("Dredging Purpose")+
+#   labs(color = "Dredging?")
+
+e<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
+         color = 'week_of_year',loadings = FALSE)+
+        ggtitle("Week ")+
   labs(color = "Dredging?")
 
-ggsave("Initial_Full_PCA.png",grid.arrange(a,b,c,d,ncol = 2),width = 10, height = 8)
+ggsave("Initial_Full_PCA.png",grid.arrange(a,b,c,e,ncol = 2),width = 10, height = 8)
 variable_contributions<- dredge_PCA$rotation
@@
                               Pool = gage_survey_cleaned$POOL,
-                              DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
                               Season = gage_survey_cleaned$Season,
@@
 
-data.frame(data_in_PC_space)|>
-  gf_point(PC2 ~ PC1, col = ~DREDGE_TYPE)
+# data.frame(data_in_PC_space)|>
+#   gf_point(PC2 ~ PC1, col = ~DREDGE_TYPE)
 
@@
 
-###xgBoost
+### xgboost og
 ```{r}
 library(xgboost)
+gage_survey_sort <- gage_survey_cleaned|>
+  arrange(DATE_START)  # If DATE_START was dropped, sort gage_survey_cleaned first
+
+gage_xgb <- gage_survey_sort |>
+  select(-lead_7, -lead_14,
+    -DATE_START,
+         -EXECYEAR)|>
+    mutate_if(is.character, factor)
+
+gage_xgb_scaled <- gage_xgb |>
+  mutate(across(where(is.numeric), ~ scale(.) |> 
+                  as.data.frame() |>
+                  pull(1), .names = "{.col}"))
+
+ctrl <- trainControl(
+  method = "timeslice",
+  initialWindow = floor(0.7 * nrow(gage_xgb_scaled)),  # 70% of data for initial training
+  horizon = 1,               # Predict 1 step ahead
+  fixedWindow = TRUE,        # Rolling window
+  verboseIter = TRUE
+)
+
+fit_dredge = caret::train(VOLUMEDREDGED ~ .,
+                 data = gage_xgb_scaled,
+                 method = "xgbTree",
+                 tuneLength = 10,  # Automatic tuning of hyperparameters
+                 trControl = ctrl
+)
+
+varImp(fit_dredge)
+xgb.plot.tree(model = fit_dredge$finalModel, trees = 0)
+
+fit_dredge
+
+import_matrix = xgb.importance(model = fit_dredge$finalModel)
+import_matrix
+xgb.plot.importance(import_matrix)
+```
+
+```{r}
+ggplot(data.frame(Observed = gage_xgb_scaled$VOLUMEDREDGED, Predicted = cv_pred), 
+       aes(x = Observed, y = Predicted)) +
+  geom_point() + 
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  labs(x = "Observed", y = "Predicted", title = "Observed vs Predicted")
+```
+
+### xgboost fast
+```{r}
+library(xgboost)
+library(dplyr)
+
+gage_survey_cleaned <- gage_survey_cleaned |>
+  arrange(DATE_START)  # Sort the data by DATE_START
+
 gage_xgb <- gage_survey_cleaned |>
-  select(-lead_7, -lead_14,-DATE_START,
-         -EXECYEAR,-DREDGE_TYPE,
-         -DREDGINGPURPOSE)|>
+  select(-lead_7, -lead_14,
+         -DATE_START,
+         -EXECYEAR,
+         -RIVER, -Season) |>
   mutate_if(is.character, factor)
 
-ctrl = trainControl(method = "cv", number = 5) 
-fit_dredge = caret::train(VOLUMEDREDGED ~ .,
-                 data = gage_xgb,
+gage_xgb_scaled <- gage_xgb |>
+  mutate(across(where(is.numeric), ~ scale(.) |> 
+                  as.data.frame() |>
+                  pull(1), .names = "{.col}"))
+
+# 80/20 split for fast test
+n <- nrow(gage_xgb_scaled)
+split_idx <- floor(0.8 * n)
+
+train_data <- gage_xgb_scaled[1:split_idx, ]
+test_data  <- gage_xgb_scaled[(split_idx + 1):n, ]
+
+train_data <- train_data %>%
+  mutate(across(where(is.factor), ~ as.integer(.)))
+
+test_data <- test_data %>%
+  mutate(across(where(is.factor), ~ as.integer(.)))
+
+# Create DMatrix
+dtrain <- xgb.DMatrix(data = as.matrix(train_data %>% select(-VOLUMEDREDGED)),
+                      label = train_data$VOLUMEDREDGED)
+dtest <- xgb.DMatrix(data = as.matrix(test_data %>% select(-VOLUMEDREDGED)))
+
+# Define quick params
+params <- list(
+  objective = "reg:squarederror",
+  eta = 0.3,                # faster learning
+  max_depth = 3,            # shallow trees for speed
+  subsample = 0.8,
+  colsample_bytree = 0.8
+)
+
+# Train with early stopping
+model <- xgb.train(
+  params = params,
+  data = dtrain,
+  nrounds = 50,             # keep small for fast testing
+  watchlist = list(train = dtrain),
+  early_stopping_rounds = 5,
+  verbose = 0
+)
+
+# Predict and evaluate quickly
+preds <- predict(model, dtest)
+actuals <- test_data$VOLUMEDREDGED
+
+# Quick RMSE
+rmse <- sqrt(mean((actuals - preds)^2))
+cat("Quick Test RMSE:", round(rmse, 3), "\n")
+
+# Feature Importance Plot
+import_matrix <- xgb.importance(model = model)
+xgb.plot.importance(import_matrix)
+
+```
+
+###Cross Validation of Model
+```{r}
+library(progress)
+pb <- progress_bar$new(total = n_outer_folds)
+set.seed(123)
+n = nrow(gage_xgb_scaled)
+n_outer_folds = 5
+
+cv_pred = vector(length = n)
+
+groups = rep(1:n_outer_folds, length = n)
+cv_groups = sample(groups, n)
+
+for(ii in 1:n_outer_folds){
+  print(Sys.time())
+  pb$tick()
+  in_test = (cv_groups == ii)
+  in_train = (cv_groups != ii)
+
+  ctrl = trainControl(method = "timeslice",
+  initialWindow = 60,   # The first 60 data points for training
+  horizon = 45,         # Forecasting horizon
+  fixedWindow = TRUE,
+  verboseIter=TRUE # Slide the window forward after each step
+)
+  fit_dredge_DCV = caret::train(VOLUMEDREDGED ~ .,
+                       data = gage_xgb_scaled[in_train, ],
                        method = "xgbTree",
-               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
-                                              max_depth = 1:10,
-                                        eta =c(0.3), gamma = 0,
-                                        colsample_bytree = 1,
-                                        min_child_weight = 1, subsample = 1),
+                     tune_grid = expand.grid(
+  nrounds = c(25, 50),        # fewer rounds
+  max_depth = c(2, 4),        # shallow trees
+  eta = c(0.3),               # fast learning rate
+  gamma = 0,                  # no regularization for speed
+  colsample_bytree = 0.8,     # reasonable feature sampling
+  min_child_weight = 1,
+  subsample = 0.8
+)
+,
                        verbosity = 0,
@@
 
-varImp(fit_dredge)
-xgb.plot.tree(model = fit_dredge$finalModel, trees = 0)
-
-fit_dredge
-
-import_matrix = xgb.importance(model = fit_dredge$finalModel)
-import_matrix
-xgb.plot.importance(import_matrix)
-
-```
-###Cross Validation of Model
-```{r}
-set.seed(123)
-n = nrow(gage_xgb)
-n_outer_folds = 5
-
-cv_pred = vector(length = n)
-
-groups = rep(1:n_outer_folds, length = n)
-cv_groups = sample(groups, n)
-
-for(ii in 1:n_outer_folds){
-  in_test = (cv_groups == ii)
-  in_train = (cv_groups != ii)
-
-  ctrl = trainControl(method = "cv", number = 5)
-  fit_dredge_DCV = caret::train(VOLUMEDREDGED ~ .,
-                       data = gage_xgb[in_train, ],
-                       method = "xgbTree",
-                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
-                                              max_depth = 1:10,
-                                        eta =c(0.3), gamma = 0,
-                                        colsample_bytree = 1,
-                                        min_child_weight = 1, subsample = 1),
-                       verbosity = 0,
-                       trControl = ctrl,
-                       na.action = na.exclude)
-
   # "Frankenstein" vector of predictions
   cv_pred[in_test] = predict(fit_dredge_DCV,
-                             newdata = gage_xgb[in_test, ])
-
-}
-
-##Og
-train<-gage_xgb[in_train, ]
-best_tune<-fit_dredge_DCV$results
-best_tune
-confusion_matrix <- confusionMatrix(cv_pred[in_test], gage_xgb[in_test, ])
-print(confusion_matrix)
-
+                             newdata = gage_xgb_scaled[in_test, ])
+  print(Sys.time())
+
+}
+
+# Post-CV model evaluation
+
+# RMSE: This is the most honest assessment of accuracy for regression
 # I'll use RMSE to assess the model, because that's the metric I used to select the model.
@@
 # RMSE
-sqrt(mean((cv_pred - gage_xgb$VOLUMEDREDGED)^2)) # This is the most honest
-                                         # assessment of accuracy
+rmse = sqrt(mean((cv_pred - gage_xgb_scaled$VOLUMEDREDGED)^2))
+cat("Root Mean Squared Error (RMSE):", rmse, "\n")
+
+# Additional Regression Metrics (MAE, R-squared, etc.)
+mae = mean(abs(cv_pred - gage_xgb_scaled$VOLUMEDREDGED))
+cat("Mean Absolute Error (MAE):", mae, "\n")
+
+# R-squared (R²) can be another good metric to evaluate model performance
+ss_total = sum((gage_xgb_scaled$VOLUMEDREDGED - mean(gage_xgb_scaled$VOLUMEDREDGED))^2)
+ss_residual = sum((gage_xgb_scaled$VOLUMEDREDGED - cv_pred)^2)
+r_squared = 1 - (ss_residual / ss_total)
+cat("R-squared:", r_squared, "\n")
+
+# Best tuning parameters from last fold
+best_tune <- fit_dredge_DCV$results
+cat("Best Hyperparameters from Last Fold:\n")
+print(best_tune)
 ```
@@
 Miss_xgb <-Miss_Cleaned |>
-  select(-lead_7, -lead_14,-DATE_START,
+  select(#-lead_7,# -lead_14,
+    -DATE_START,
          -EXECYEAR,-DREDGE_TYPE,
@@
 
-ctrl = trainControl(method = "cv", number = 10) 
+ctrl = trainControl(method = "timeslice",
+                      initialWindow = 60,   # The first 60 data points for training
+                      horizon = 45,         # Forecasting horizon
+                      fixedWindow = TRUE,   # Slide the window forward after each step
+                      verboseIter = TRUE) 
 Miss_xgb_fit = caret::train(VOLUMEDREDGED ~ .,
@@
 library(keras)
+
+# Function to create sequences based on window size
+create_sequences <- function(data, target_column, sequence_length) {
+  X <- list()
+  y <- list()
+
+  for (i in (sequence_length + 1):nrow(data)) {
+    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])  # Features excluding target
+    y[[i - sequence_length]] <- data[i, target_column]  # Target column
+  }
+
+  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))  # Reshape to 3D array
+  y <- unlist(y)
+
+  return(list(X = X, y = y))
+}
+
+# Example: Create sequences for 7-day and 14-day windows
+window_size_7 <- 7
+window_size_14 <- 14
+
+# Load necessary libraries
+library(recipes)
+
+# Assume gage_xgb is your original dataset
+gage_xgb_scaled_recipe <- recipe(~ ., data = gage_xgb) |>
+  step_dummy(all_nominal(), one_hot = TRUE) |>
+  step_scale(all_numeric_predictors()) |>
+  step_center(all_numeric_predictors())
+
+# Prep the recipe
+gage_xgb_prepped <- prep(gage_xgb_scaled_recipe, training = gage_xgb)
+
+# Apply the scaling to the data
+gage_xgb_scaled <- bake(gage_xgb_prepped, new_data = gage_xgb)
+
+# View scaled data
+head(gage_xgb_scaled)
+
+gage_xgb_scaled_clean <- gage_xgb_scaled|>
+  select(-lead_7,-lead_14)
+
+# Assume `gage_xgb_scaled` is your scaled dataset and the target column is 'VOLUMEDREDGED'
+data_windows_7 <- create_sequences(gage_xgb_scaled_clean, target_column = which(names(gage_xgb_scaled_clean) == "VOLUMEDREDGED"), sequence_length = window_size_7)
+data_windows_14 <- create_sequences(gage_xgb_scaled_clean, target_column = which(names(gage_xgb_scaled_clean) == "VOLUMEDREDGED"), sequence_length = window_size_14)
+
+# X_train and y_train for 7-day window
+X_train_7 <- data_windows_7$X
+y_train_7 <- data_windows_7$y
+
+# X_train and y_train for 14-day window
+X_train_14 <- data_windows_14$X
+y_train_14 <- data_windows_14$y
+
+# Define the train-test split
+train_size_7 <- floor(0.8 * length(y_train_7))  # 80% for training
+train_size_14 <- floor(0.8 * length(y_train_14))  # 80% for training
+
+# Training and Testing Data for 7-day window
+X_train_7_final <- X_train_7[1:train_size_7,,]
+y_train_7_final <- y_train_7[1:train_size_7]
+X_test_7_final <- X_train_7[(train_size_7+1):length(y_train_7),,]
+y_test_7_final <- y_train_7[(train_size_7+1):length(y_train_7)]
+
+# Training and Testing Data for 14-day window
+X_train_14_final <- X_train_14[1:train_size_14,,]
+y_train_14_final <- y_train_14[1:train_size_14]
+X_test_14_final <- X_train_14[(train_size_14+1):length(y_train_14),,]
+y_test_14_final <- y_train_14[(train_size_14+1):length(y_train_14)]
+
+# Build the model
 model_7 <- keras_model_sequential() |>
-  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(7, ncol(gage_xgb) - 1))  |>
+  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(7, ncol(gage_xgb_scaled_clean) - 1))  |>
   layer_dropout(rate = 0.2)|>
@@
 
+# model_7 %>% compile(
+#   loss = 'mean_squared_error',
+#   optimizer = 'adam'
+# )
+
 model_7 %>% compile(
   loss = 'mean_squared_error',
-  optimizer = 'adam'
-)
-
-# Function to create sequences based on window size
+  optimizer = optimizer_adam(learning_rate = 0.001)  # Try also 0.0005 or 0.0001
+)
+
+callback <- callback_early_stopping(monitor = "val_loss", patience = 10, restore_best_weights = TRUE)
+
+history_7 <- model_7 |> fit(
+  X_train_7_final, 
+  y_train_7_final,
+  epochs = 1000,
+  batch_size = 32,
+  validation_data = list(X_test_7_final, y_test_7_final),
+  callbacks = list(callback)
+)
+
+# Train the model for the 7-day window
+# history_7 <- model_7 %>% fit(
+#   X_train_7_final, 
+#   y_train_7_final,
+#   epochs = 1000, 
+#   batch_size = 32, 
+#   validation_data = list(X_test_7_final, y_test_7_final)
+# )
+
+# Evaluate model on 7-day window
+score_7 <- model_7 %>% evaluate(X_test_7_final, y_test_7_final)
+cat('Test loss for 7-day window:', score_7, "\n")
+
+predictions_7 <- model_7 %>% predict(X_test_7_final)
+
+```
+
+### LSTM method2
+```{r}
+library(tensorflow)
+library(keras)
+library(recipes)
+library(dplyr)
+
+# Set window sizes
+window_size_7 <- 7
+window_size_14 <- 14
+
+# --- Step 1: Separate and scale features & target ---
+target_col <- "VOLUMEDREDGED"
+gage_target <- gage_xgb[[target_col]]
+gage_features <- gage_xgb |> select(-all_of(target_col))
+
+# Scale predictors
+gage_recipe <- recipe(~ ., data = gage_features) |>
+  step_dummy(all_nominal(), one_hot = TRUE) |>
+  step_scale(all_numeric_predictors()) |>
+  step_center(all_numeric_predictors())
+gage_recipe_prep <- prep(gage_recipe, training = gage_features)
+gage_features_scaled <- bake(gage_recipe_prep, new_data = gage_features)
+
+# Scale target manually
+target_mean <- mean(gage_target, na.rm = TRUE)
+target_sd <- sd(gage_target, na.rm = TRUE)
+gage_target_scaled <- (gage_target - target_mean) / target_sd
+
+# Recombine scaled features and target
+gage_scaled_clean <- bind_cols(gage_features_scaled, VOLUMEDREDGED = gage_target_scaled)
+
+# Optionally remove future-looking targets
+gage_scaled_clean <- gage_scaled_clean |> select(-lead_7, -lead_14)
+
+# --- Step 2: Sequence creation function ---
 create_sequences <- function(data, target_column, sequence_length) {
@@
   for (i in (sequence_length + 1):nrow(data)) {
-    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])  # Features excluding target
-    y[[i - sequence_length]] <- data[i, target_column]  # Target column
-  }
-
-  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))  # Reshape to 3D array
+    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])
+    y[[i - sequence_length]] <- data[i, target_column]
+  }
+
+  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))
   y <- unlist(y)
@@
 
-# Example: Create sequences for 7-day and 14-day windows
-window_size_7 <- 7
-window_size_14 <- 14
-
-# Load necessary libraries
+# --- Step 3: Create sequences ---
+target_index <- which(names(gage_scaled_clean) == "VOLUMEDREDGED")
+data_windows_7 <- create_sequences(gage_scaled_clean, target_column = target_index, sequence_length = window_size_7)
+
+X <- data_windows_7$X
+y <- data_windows_7$y
+
+# Split into train/test
+train_size <- floor(0.8 * length(y))
+X_train <- X[1:train_size,,]
+y_train <- y[1:train_size]
+X_test <- X[(train_size+1):length(y),,]
+y_test <- y[(train_size+1):length(y)]
+
+# --- Step 4: Build model ---
+model <- keras_model_sequential() |>
+  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(window_size_7, ncol(gage_scaled_clean) - 1)) |>
+  layer_dropout(rate = 0.2) |>
+  layer_lstm(units = 50, return_sequences = TRUE) |>
+  layer_dropout(rate = 0.2) |>
+  layer_lstm(units = 50) |>
+  layer_dropout(rate = 0.2) |>
+  layer_dense(units = 1)
+
+model %>% compile(
+  loss = 'mean_squared_error',
+  optimizer = optimizer_adam(learning_rate = 0.001)
+)
+
+callback <- callback_early_stopping(monitor = "val_loss", patience = 10, restore_best_weights = TRUE)
+
+history <- model %>% fit(
+  X_train, y_train,
+  epochs = 1000,
+  batch_size = 32,
+  validation_data = list(X_test, y_test),
+  callbacks = list(callback)
+)
+
+# --- Step 5: Predict and invert scaling ---
+pred_scaled <- model %>% predict(X_test)
+pred <- pred_scaled * target_sd + target_mean
+actual <- y_test * target_sd + target_mean
+
+# --- Step 6: Evaluate on original scale ---
+rmse <- sqrt(mean((actual - pred)^2))
+r_squared <- 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)
+mae <- mean(abs(actual - pred))
+
+cat("RMSE:", rmse, "\n")
+cat("R-squared:", r_squared, "\n")
+cat("MAE:", mae, "\n")
+```
+
+##Lets try predicting a binary yes/no for dredging
+```{r}
+library(tensorflow)
+library(keras)
 library(recipes)
-
-# Assume gage_xgb is your original dataset
-gage_xgb_scaled_recipe <- recipe(~ ., data = gage_xgb) %>%
-  # Define step to scale all numeric predictors
-  step_scale(all_numeric_predictors()) %>%
-  # You can also center them if needed
+library(dplyr)
+
+# --- Step 1: Parameters ---
+window_size <- 7
+forecast_horizon <- 7  # Predict if dredging will occur in the next 7 days
+target_col <- "VOLUMEDREDGED"
+
+# --- Step 2: Feature/target separation and scaling ---
+gage_target <- gage_xgb[[target_col]]
+gage_features <- gage_xgb |> select(-all_of(target_col))
+
+# Use recipes to scale and encode predictors
+gage_recipe <- recipe(~ ., data = gage_features) |>
+  step_dummy(all_nominal(), one_hot = TRUE) |>
+  step_scale(all_numeric_predictors()) |>
   step_center(all_numeric_predictors())
 
-# Prep the recipe
-gage_xgb_prepped <- prep(gage_xgb_scaled_recipe, training = gage_xgb)
-
-# Apply the scaling to the data
-gage_xgb_scaled <- bake(gage_xgb_prepped, new_data = gage_xgb)
-
-# View scaled data
-head(gage_xgb_scaled)
-
-# Assume `gage_xgb_scaled` is your scaled dataset and the target column is 'VOLUMEDREDGED'
-data_windows_7 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_7)
-data_windows_14 <- create_sequences(gage_xgb_scaled, target_column = which(names(gage_xgb) == "VOLUMEDREDGED"), sequence_length = window_size_14)
-
-# X_train and y_train for 7-day window
-X_train_7 <- data_windows_7$X
-y_train_7 <- data_windows_7$y
-
-# X_train and y_train for 14-day window
-X_train_14 <- data_windows_14$X
-y_train_14 <- data_windows_14$y
-
-# Define the train-test split
-train_size_7 <- floor(0.8 * length(y_train_7))  # 80% for training
-train_size_14 <- floor(0.8 * length(y_train_14))  # 80% for training
-
-# Training and Testing Data for 7-day window
-X_train_7_final <- X_train_7[1:train_size_7,,]
-y_train_7_final <- y_train_7[1:train_size_7]
-X_test_7_final <- X_train_7[(train_size_7+1):length(y_train_7),,]
-y_test_7_final <- y_train_7[(train_size_7+1):length(y_train_7)]
-
-# Training and Testing Data for 14-day window
-X_train_14_final <- X_train_14[1:train_size_14,,]
-y_train_14_final <- y_train_14[1:train_size_14]
-X_test_14_final <- X_train_14[(train_size_14+1):length(y_train_14),,]
-y_test_14_final <- y_train_14[(train_size_14+1):length(y_train_14)]
-
-# Train the model for the 7-day window
-history_7 <- model_7 %>% fit(
-  X_train_7_final, 
-  y_train_7_final,
-  epochs = 20, 
+gage_recipe_prep <- prep(gage_recipe, training = gage_features)
+gage_features_scaled <- bake(gage_recipe_prep, new_data = gage_features)
+
+# Keep target as-is (not scaled), used for binary labeling
+gage_scaled_clean <- bind_cols(gage_features_scaled, VOLUMEDREDGED = gage_target)
+
+# Optional: remove future-looking columns
+gage_scaled_clean <- gage_scaled_clean |> select(-lead_7, -lead_14)
+
+# --- Step 3: Sequence creation for classification ---
+create_binary_sequences <- function(data, target_column, sequence_length, forecast_horizon) {
+  X <- list()
+  y <- list()
+
+  for (i in (sequence_length + 1):(nrow(data) - forecast_horizon)) {
+    X[[i - sequence_length]] <- as.matrix(data[(i - sequence_length):(i - 1), -target_column])
+
+    future_window <- data[i:(i + forecast_horizon - 1), target_column]
+    dredging_needed <- as.integer(any(future_window > 0))
+
+    y[[i - sequence_length]] <- dredging_needed
+  }
+
+  X <- array(unlist(X), dim = c(length(X), sequence_length, ncol(data) - 1))
+  y <- unlist(y)
+
+  return(list(X = X, y = y))
+}
+
+# --- Step 4: Generate data ---
+target_index <- which(names(gage_scaled_clean) == "VOLUMEDREDGED")
+data_windows <- create_binary_sequences(
+  gage_scaled_clean,
+  target_column = target_index,
+  sequence_length = window_size,
+  forecast_horizon = forecast_horizon
+)
+
+X <- data_windows$X
+y <- data_windows$y
+
+# --- Step 5: Train/test split ---
+train_size <- floor(0.8 * length(y))
+X_train <- X[1:train_size,,]
+y_train <- y[1:train_size]
+X_test <- X[(train_size + 1):length(y),,]
+y_test <- y[(train_size + 1):length(y)]
+
+# --- Step 6: LSTM model ---
+model <- keras_model_sequential() |>
+  layer_lstm(units = 50, return_sequences = TRUE, input_shape = c(window_size, dim(X)[3])) |>
+  layer_dropout(rate = 0.2) |>
+  layer_lstm(units = 50, return_sequences = TRUE) |>
+  layer_dropout(rate = 0.2) |>
+  layer_lstm(units = 50) |>
+  layer_dropout(rate = 0.2) |>
+  layer_dense(units = 1, activation = "sigmoid")  # Binary output
+
+model %>% compile(
+  loss = 'binary_crossentropy',
+  optimizer = optimizer_adam(learning_rate = 0.001),
+  metrics = c('accuracy', metric_auc())
+)
+
+# --- Step 7: Train model ---
+callback <- callback_early_stopping(monitor = "val_loss", patience = 10, restore_best_weights = TRUE)
+
+history <- model %>% fit(
+  X_train, y_train,
+  epochs = 100,
   batch_size = 32,
-  validation_data = list(X_test_7_final, y_test_7_final)
-)
-
-# Evaluate model on 7-day window
-score_7 <- model_7 %>% evaluate(X_test_7_final, y_test_7_final)
-cat('Test loss for 7-day window:', score_7, "\n")
-
-predictions_7 <- model_7 %>% predict(X_test_7_final)
-```
+  validation_data = list(X_test, y_test),
+  callbacks = list(callback)
+)
+
+# --- Step 8: Evaluate and predict ---
+pred_probs <- model %>% predict(X_test)
+pred_classes <- ifelse(pred_probs > 0.5, 1, 0)
+
+# Confusion matrix
+confusion_matrix <- table(Predicted = pred_classes, Actual = y_test)
+print(confusion_matrix)
+
+# Metrics
+accuracy <- mean(pred_classes == y_test)
+TP <- sum(pred_classes == 1 & y_test == 1)
+FP <- sum(pred_classes == 1 & y_test == 0)
+FN <- sum(pred_classes == 0 & y_test == 1)
+
+precision <- ifelse((TP + FP) > 0, TP / (TP + FP), 0)
+recall <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)
+f1 <- ifelse((precision + recall) > 0, 2 * precision * recall / (precision + recall), 0)
+
+cat("Accuracy:", accuracy, "\n")
+cat("Precision:", precision)
+
+```
+
+```{r}
+library(tensorflow)
+library(keras)
+library(recipes)
+library(dplyr)
+
+prepare_lstm_data <- function(data) {
+  # Select relevant features
+  relevant_features <- c(
+    "VOLUMEDREDGED", "UP_RIV_MIL", "DN_RIV_MIL", "POOL", "DATE_START",
+    names(select(data, matches("MI\\d+")))
+  )
+
+  data_subset <- data %>%
+    select(all_of(relevant_features)) %>%
+    arrange(DATE_START)
+
+  # Convert POOL to factor and then numeric
+  data_subset$POOL <- as.numeric(factor(data_subset$POOL))
+
+  # Create binary target
+  data_subset$dredge_needed <- as.integer(data_subset$VOLUMEDREDGED > 0)
+
+  # Remove DATE_START after arranging
+  data_subset <- data_subset %>%
+    select(-DATE_START)
+
+  # Scale numeric features
+  recipe_obj <- recipe(~ ., data = data_subset) %>%
+    step_normalize(all_numeric_predictors())
+
+  prepped_recipe <- prep(recipe_obj)
+  scaled_data <- bake(prepped_recipe, new_data = data_subset)
+
+  return(scaled_data)
+}
+
+# 2. Modified sequence creation
+create_sequences <- function(data, window_size = 45) {
+  # Ensure all columns are numeric
+  numeric_data <- data %>%
+    mutate(across(everything(), as.numeric))
+
+  X <- list()
+  y <- list()
+
+  for(i in 1:(nrow(numeric_data) - window_size)) {
+    # Exclude dredge_needed column from features
+    X[[i]] <- as.matrix(numeric_data[i:(i + window_size - 1), 
+                                   !names(numeric_data) %in% c("dredge_needed")])
+    y[[i]] <- numeric_data$dredge_needed[i + window_size]
+  }
+
+  X <- array(unlist(X), 
+            dim = c(length(X), 
+                   window_size, 
+                   ncol(numeric_data) - 1))  # -1 for dredge_needed column
+  y <- unlist(y)
+
+  return(list(X = X, y = y))
+}
+
+# 3. Main execution
+# Prepare data
+scaled_data <- prepare_lstm_data(Miss_Cleaned)
+
+# Create sequences
+sequences <- create_sequences(scaled_data)
+
+# Split data
+split_idx <- floor(0.8 * length(sequences$y))
+X_train <- sequences$X[1:split_idx,,]
+y_train <- sequences$y[1:split_idx]
+X_val <- sequences$X[(split_idx+1):length(sequences$y),,]
+y_val <- sequences$y[(split_idx+1):length(sequences$y)]
+
+# Verify data types
+print("Training data dimensions:")
+print(dim(X_train))
+print("Training labels length:")
+print(length(y_train))
+print("First few values of X_train:")
+print(head(X_train[1,,]))
+print("First few values of y_train:")
+print(head(y_train))
+
+# Train model
+model_results <- train_lstm_model(X_train, y_train, X_val, y_val)
+
+# Evaluate model
+predictions <- predict(model_results$model, X_val)
+binary_predictions <- ifelse(predictions > 0.5, 1, 0)
+
+# Calculate metrics
+confusion_matrix <- table(Actual = y_val, Predicted = binary_predictions)
+accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
+recall <- confusion_matrix[2,2] / sum(confusion_matrix[2,])
+
+print("Confusion Matrix:")
+print(confusion_matrix)
+print(paste("Accuracy:", round(accuracy, 3)))
+print(paste("Recall:", round(recall, 3)))
+
+# First, let's check the distribution of actual values
+print("Distribution of actual values (y_val):")
+print(table(y_val))
+
+# Check the raw predictions before binary conversion
+print("Summary of raw predictions:")
+print(summary(predictions))
+
+# Modified evaluation code with error handling
+evaluate_binary_predictions <- function(y_true, y_pred) {
+  # Convert predictions to binary
+  binary_pred <- ifelse(y_pred > 0.5, 1, 0)
+
+  # Create confusion matrix
+  conf_matrix <- table(Actual = y_true, Predicted = binary_pred)
+  print("Confusion Matrix:")
+  print(conf_matrix)
+
+  # Calculate metrics with error handling
+  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
+
+  # Handle case where there are no positive predictions
+  if(length(conf_matrix) == 1 || !any(binary_pred == 1)) {
+    recall <- 0
+    precision <- 0
+    f1 <- 0
+  } else {
+    TP <- conf_matrix["1", "1"]
+    FN <- conf_matrix["1", "0"]
+    FP <- conf_matrix["0", "1"]
+
+    recall <- ifelse(TP + FN > 0, TP / (TP + FN), 0)
+    precision <- ifelse(TP + FP > 0, TP / (TP + FP), 0)
+    f1 <- ifelse(precision + recall > 0, 
+                 2 * (precision * recall) / (precision + recall), 
+                 0)
+  }
+
+  return(list(
+    confusion_matrix = conf_matrix,
+    accuracy = accuracy,
+    recall = recall,
+    precision = precision,
+    f1 = f1
+  ))
+}
+
+# Modify the model to address class imbalance
+train_lstm_model <- function(X_train, y_train, X_val, y_val) {
+  # Calculate class weights
+  n_neg <- sum(y_train == 0)
+  n_pos <- sum(y_train == 1)
+  total <- n_neg + n_pos
+
+  class_weight <- list(
+    "0" = 1,
+    "1" = n_neg/n_pos  # Give more weight to minority class
+  )
+
+  model <- keras_model_sequential() %>%
+    layer_lstm(units = 64, return_sequences = TRUE, 
+              input_shape = c(dim(X_train)[2], dim(X_train)[3])) %>%
+    layer_dropout(0.2) %>%
+    layer_lstm(units = 32) %>%
+    layer_dropout(0.2) %>%
+    layer_dense(units = 16, activation = "relu") %>%
+    layer_dense(units = 1, activation = "sigmoid")
+
+  model %>% compile(
+    loss = "binary_crossentropy",
+    optimizer = optimizer_adam(learning_rate = 0.001),
+    metrics = c("accuracy", "AUC")
+  )
+
+  history <- model %>% fit(
+    X_train, y_train,
+    epochs = 50,
+    batch_size = 32,
+    validation_data = list(X_val, y_val),
+    class_weight = class_weight,  # Add class weights
+    callbacks = list(
+      callback_early_stopping(
+        monitor = "val_loss",
+        patience = 10,
+        restore_best_weights = TRUE
+      )
+    )
+  )
+
+  return(list(model = model, history = history))
+}
+
+# Retrain model with class weights
+model_results <- train_lstm_model(X_train, y_train, X_val, y_val)
+
+# Make predictions
+predictions <- predict(model_results$model, X_val)
+
+# Evaluate with new function
+results <- evaluate_binary_predictions(y_val, predictions)
+
+# Print detailed results
+print("Evaluation Results:")
+print(paste("Accuracy:", round(results$accuracy, 3)))
+print(paste("Recall:", round(results$recall, 3)))
+print(paste("Precision:", round(results$precision, 3)))
+print(paste("F1 Score:", round(results$f1, 3)))
+
+# Plot training history
+plot(model_results$history)
+
+# Additional analysis of predictions
+hist(predictions, 
+     main = "Distribution of Predictions",
+     xlab = "Predicted Probability",
+     breaks = 20)
+
+```

### 2025-06-26 CSAT_Dredge_Data.R
reviewing CSAT derived dredge data

@@ -1,0 +1,109 @@
+library(readr)
+library(dplyr)
+library(stringr)
+library(purrr)
+library(ggplot2)
+library(lubridate)
+library(tidyverse)
+library(zoo)
+options(scipen = 100, digits = 2)
+dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
+# Set folder path
+folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101__to__20241231"
+
+reach_data<-read_csv("C:/workspace/DRAFT/ReachData.csv")
+
+# List matching files (ends with dredged_date_vol.csv)
+file_list <- list.files(path = folder_path,
+                        pattern = "_dredged_date_vol\\.csv$",
+                        full.names = TRUE)
+
+# Function to extract metadata and read file
+read_and_annotate <- function(file_path) {
+  # Extract file name only
+  file_name <- basename(file_path)
+  
+  # Use regex to extract River, Pool, and Reach
+  # Pattern: CEMVR_(River)_(Pool)_MVR_(Reach)_dredged_date_vol
+  matches <- str_match(file_name, "^CEMVR_([A-Z]{2})_([A-Z0-9]+)_MVR_([0-9]+)_dredged_date_vol\\.csv$")
+  
+  if (is.na(matches[1,1])) {
+    warning(paste("Filename does not match expected pattern:", file_name))
+    return(NULL)
+  }
+
+  
+  reachname <- 
+  river <- matches[1,2]
+  pool <- matches[1,3]
+  reach <- matches[1,4]
+  reachname <-paste0("CEMVR_",river,"_",pool,"_MVR_",reach) 
+  
+  # Read data
+  df <- read_csv(file_path, show_col_types = FALSE)
+  
+  # Add extracted info
+  df <- df %>%
+    mutate(channelreachidpk = reachname,
+           River = river,
+           Pool = pool,
+           Reach = reach)
+  
+  return(df)
+}
+
+# Apply to all files and combine
+csat_dredge_data <- file_list %>%
+  lapply(read_and_annotate) %>%
+  bind_rows()
+
+csat_dredged<-rename(csat_dredge_data, VolumeChange = `Volume Change (cy)`)
+
+glimpse(csat_dredged)
+
+data_cleaned <- csat_dredged|>
+  filter(VolumeChange>500)
+
+
[…]

### 2025-06-26 CSAT_Rates.Rmd
CSAT rate wrangling and spot checks

---
title: "CSAT_Wrangling_ShoalingRates"
output: html_document
date: "2025-05-12"
---

```{r}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(zoo)
```

```{r}
# Set folder path
folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101__to__20241231"

reach_data<-read_csv("C:/workspace/DRAFT/ReachData.csv")

```

```{r, message=FALSE}
### Shoaling Rates

files <-list.files(path = folder_path,
                                 pattern = "_SurveyPairVolumeDifference\\.csv$",
                                 full.names = TRUE)

read_and_label2 <- function(file_path) {
  # Extract filename
  file_name <- basename(file_path)

  # Remove extension
  file_name_no_ext <- tools::file_path_sans_ext(file_name)

  # Split by underscore
  parts <- str_split(file_name_no_ext, "_", simplify = TRUE)

  # Extract relevant parts
  river <- parts[1]
  pool <- parts[2]
  reach <- parts[4]  # MVR is at [3], 14 or 22 is at [4]

  # Read the data
  data <- read_csv(file_path, show_col_types = FALSE)

  # Add new columns
  data <- data %>%
    mutate(
      river = river,
      pool = pool,
      reach = reach,
      source_file = file_name
    )

  return(data)
}

# Apply the function to all files and bind into one big dataframe
combined_data <- map_dfr(files, read_and_label2)
```

```{r}
# Check it out
glimpse(combined_data)

combined_data_clean<-combined_data|>
  mutate(Before_date = ymd(SurveyDateBefore),
         After_date = ymd(SurveyDateAfter))|>
  mutate(DateDiff = as.numeric(After_date - Before_date))|>
  mutate(DailyShoalingRate_ftperday = AnnualShoalingRate_ftperyr/DateDiff,
         DailyShoalingVolume_Cyperday = AnnualShoalingVolume_CYperyr/DateDiff,
         VolChange_Cyperday = NetVolumeChange_CY/DateDiff)
#  filter(DailyShoalingVolume_Cyperday < 12500000 & DailyShoalingRate_ftperday < 900)
```

###Plot Rates by day
```{r}
ggplot(combined_data_clean, aes(x = After_date, y = AnnualShoalingVolume_CYperyr, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(-2000000,1000000)+
  facet_wrap(~ river)

ggplot(combined_data_clean, aes(x = After_date, y = AnnualShoalingRate_ftperyr, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(NA, 275)+
  facet_wrap(~ river)

ggplot(combined_data_clean, aes(x = After_date, y = NetVolumeChange_CY, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(NA, 275)+
  facet_wrap(~ river)

ggplot(combined_data_clean, aes(x = Before_date, y = NetVolumeChange_CY)) +
  geom_point() +
  theme(legend.position = "bottom")+
  geom_line(alpha = 0.5) +
  labs(title = "Net Volume Change in Sediment (CY) Over Time",
       x = "Date of Survey After",
       y = "Net Volume Change (CY)",
       color = "River") +
  theme_minimal() +
    facet_wrap(~ pool, scales = "free_y")

```
Let's only look at rate differences calculated within a year 
```{r}
combined_data_year <- combined_data_clean|>
  mutate(yr_diff = abs(year(After_date) - year(Before_date)))|>
  filter(yr_diff <1)|>
  filter(pool != "AL")|>
  filter(pool != "CS")
```

```{r}
ggplot(combined_data_year, aes(x = After_date, y = AnnualShoalingVolume_CYperyr, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(NA,2000000)+
  facet_wrap(~ river)

ggplot(combined_data_year, aes(x = After_date, y = AnnualShoalingRate_ftperyr, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(NA, 275)+
  facet_wrap(~ river)

ggplot(combined_data_year, aes(x = After_date, y = NetVolumeChange_CY, colour = river)) +
  geom_point() + 
  geom_line()+
#  ylim(NA, 275)+
  facet_wrap(~ river)

ggplot(combined_data_year, aes(x = Before_date, y = NetVolumeChange_CY, color = river)) +
  geom_point() +
  theme(legend.position = "bottom")+
  geom_line(alpha = 0.5) +
  labs(title = "Net Volume Change in Sediment (CY) Over Time By Pool",
       x = "Survey Date",
       y = "Net Volume Change (CY)",
       color = "River") +
  theme_minimal() +
    facet_wrap(~ pool, scales = "free_y")
```

### different view
```{r}
daily_expanded <- combined_data_clean |>
  # Create a sequence of dates for each row
  rowwise() |>
  mutate(DateSeq = list(seq(Before_date, After_date, by = "day"))) |>
  ungroup() |>
  # Expand to one row per day
  unnest(cols = c(DateSeq)) |>
  rename(Date = DateSeq) |>
  # Keep only the relevant daily rate values
  select(Date, DailyShoalingVolume_Cyperday,DailyShoalingRate_ftperday, river, pool, reach)

write_csv(daily_expanded, "C:/workspace/DRAFT/CSAT_Data/SurveyPairVolumeDifference_PreRollingAvg.csv")

  glimpse(daily_expanded)

# daily_series <- daily_expanded |>
#   group_by(Date, river, pool, reach) |>
#   summarise(DailyVolume = sum(DailyShoalingVolume_Cyperday, na.rm = TRUE), 
#             DailyRate = sum(DailyShoalingRate_ftperday, na.rm = TRUE),
#             .groups = "drop") |>
#   filter(pool == "LA")|>
#   arrange(river, pool, reach, Date) |>
#   group_by(river, pool, reach) |>
#   summarise(mutate(RollingAvgVol_7d = rollmean(DailyVolume, k = 7, fill = NA, align = "right"),
#          RollingAvgRate_7d = rollmean(DailyRate, k = 7, fill = NA, align = "right")))  |>
#   ungroup()

daily_avg <- daily_expanded %>%
  group_by(Date, river, pool, reach) %>%
  summarise(
    avg_volume = mean(DailyShoalingVolume_Cyperday, na.rm = TRUE),
    count_volume = sum(!is.na(DailyShoalingVolume_Cyperday)),
     sd_volume = sd(DailyShoalingVolume_Cyperday, na.rm = TRUE),
    se_volume = sd_volume/sqrt(n()),
    avg_rate = mean(DailyShoalingRate_ftperday, na.rm = TRUE),
    count_rate = sum(!is.na(DailyShoalingRate_ftperday)),
    sd_rate = sd(DailyShoalingRate_Cyperday, na.rm = TRUE),
    se_rate = sd_rate/sqrt(n()),
    .groups = "drop"
  )|>
  mutate(site = paste(river, pool, reach, sep = "_"))

daily_avg_rolled <- daily_avg %>%
  arrange(site, Date) %>%
  group_by(site) %>%
  mutate(
    avg_volume_7day = rollmean(avg_volume, k = 7, fill = NA, align = "right"),
     avg_volume_7day_upper = avg_volume_7day + 1.96 * rollmean(se_volume, k = 7, fill = NA),
    avg_volume_7day_lower = avg_volume_7day - 1.96 * rollmean(se_volume, k = 7, fill = NA),
    avg_rate_7day = rollmean(avg_rate, k = 7, fill = NA, align = "right"),
  ) %>%
  ungroup()

write_csv(daily_avg_rolled, "C:/workspace/DRAFT/CSAT_Data/SurveyPairVolumeDifference.csv")

# Loop through each unique value in the `pool` column
for (pool_value in unique(daily_avg_rolled$pool)) {

# Create a new dataframe for each pool value, named dynamically
  assign(paste0("P_", pool_value), daily_avg_rolled %>% filter(pool == pool_value))
}

year_plots<-year_plots |>
  mutate(Date_norm = as.Date("2000-01-01") + as.numeric(Date - lubridate::floor_date(Date, "year")))

year_plots |>
  ggplot(aes(Date_norm, avg_rate_7day, color = factor(year))) +
  geom_line() +
  scale_x_date(date_labels = "%b")+
  facet_wrap(~reach,scales = "free_y")+
  ggtitle("Pool 18 avg 7 day volumes across each reach")

daily_avg_rolled|>
  filter(river == "IL")|>
  ggplot(aes(x = Date, y = avg_volume_7day, colour = river)) +
  geom_point() + 
  ylim(NA,125000)+
  geom_line()

daily_avg_rolled|>
  filter(river == "UM")|>
  ggplot(aes(x = Date, y = avg_volume_7day, colour = river)) +
  geom_point() + 
  ylim(NA,125000)+
  geom_line()

daily_avg_rolled |>
  filter(river == "IL")|>
  ggplot(aes(x = Date, y = count_rate)) +
  geom_col(fill = "dodgerblue", alpha = 0.60)+
  facet_wrap(~pool)

daily_avg_rolled |>
  filter(river == "UM")|>
  ggplot(aes(x = Date, y = count_rate)) +
  geom_col(fill = "dodgerblue", alpha = 0.60)+
  facet_wrap(~pool)

#, scales = "free_y"

# Then: pivot to wide format
daily_volume_wide <- daily_avg %>%
  select(Date, site, avg_volume) %>%
  pivot_wider(
    names_from = site,
    values_from = avg_volume
  )|>
  mutate(year = year(Date))

daily_rate_wide <- daily_avg %>%
  select(Date, site, avg_rate) %>%
  pivot_wider(
    names_from = site,
    values_from = avg_rate
  )

daily_expanded |>
  filter(pool == "20")|>
  filter(year(Date) >= 2015 & year(Date) <= 2020)|>
  ggplot(aes(x = Date, y = DailyShoalingVolume_Cyperday, colour = river)) +
  geom_point() + 
  geom_line()+
  facet_wrap(~reach)

```

```{r}
# For the yearly comparison plot
year_plots |>
  ggplot(aes(Date_norm, avg_rate_7day, color = factor(year))) +
  geom_line() +
  scale_x_date(date_labels = "%b") +
  geom_ribbon(aes(ymin = avg_rate_7day_lower, ymax = avg_rate_7day_upper, fill = factor(year)), alpha = 0.2) +
  facet_wrap(~reach, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Pool 18 Average 7-day Volumes Across Each Reach",
    x = "Month",
    y = "7-day Average Rate",
    color = "Year"
  )

# For the river-specific plots
daily_avg_rolled |>
  filter(river == "IL") |>
  ggplot(aes(x = Date, y = avg_volume_7day)) +
  geom_line(color = "blue") +
   geom_ribbon(aes(ymin = avg_volume_7day_lower, ymax = avg_volume_7day_upper), 
              fill = "blue", alpha = 0.2) +
  ylim(NA, 125000) +
  theme_minimal() +
  labs(
    title = "Illinois River 7-day Average Volumes",
    x = "Date",
    y = "Volume (CY)"
  )
```

### 2025-09-03 CSAT_Rates.Rmd
catching up for capstone

@@
 # Set folder path
-folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101__to__20241231"
+folder_path <- "C:/workspace/CSAT/CSAT_distribution/output/CEMVR/19990101_to_20241231_Aug"
 
 reach_data<-read_csv("C:/workspace/DRAFT/ReachData.csv")
+
+```
+
+```{r}
+surveyUpdate_table<-read_delim("SurveyUpdate.txt")
+dredge_data <-  read.csv("Dredge_Event_data.csv")
+
+dredge_data<- dredge_data |>
+  filter(!is.na(POSTHYDROSURVEY_ID))|>
+  filter(POSTHYDROSURVEY_ID != "")
+
+dredge_survey_joined <- left_join(surveyUpdate_table, dredge_data, 
+                                  by = join_by(SurveyID == POSTHYDROSURVEY_ID), 
+                                  relationship = "many-to-many", keep = TRUE)|>
+  mutate(SurveyType, case_when(
+    !is.na(POSTHYDROSURVEY_ID) ~ "AD",
+    TRUE ~ SurveyType))
 
@@
 
+bad_data <- combined_data|>
+  filter(abs(AnnualShoalingRate_ftperyr) >= 30)
+
 combined_data_clean<-combined_data|>
@@
          VolChange_Cyperday = NetVolumeChange_CY/DateDiff)
-#  filter(DailyShoalingVolume_Cyperday < 12500000 & DailyShoalingRate_ftperday < 900)
 ```
@@
 ```
+
 Let's only look at rate differences calculated within a year 
@@
 ```{r}
-ggplot(combined_data_year, aes(x = After_date, y = AnnualShoalingVolume_CYperyr, colour = river)) +
+outliers_zscore <- combined_data_clean |>
+  group_by(pool) |>
+  mutate(score = as.numeric(scale(AnnualShoalingRate_ftperyr))) |>
+  filter(abs(score) > 3) |>  # Points more than 3 standard deviations from mean
+  arrange(pool, desc(abs(score)))|>
+  filter(abs(AnnualShoalingRate_ftperyr) >= 30)
+
+write_csv(outliers_zscore,"Outliers_shoalingRateOmitted.csv")
+```
+
+```{r}
+
+ggplot(combined_data_clean, aes(x = After_date, y = AnnualShoalingVolume_CYperyr, colour = river)) +
   geom_point() + 
@@
 
-ggplot(combined_data_year, aes(x = After_date, y = AnnualShoalingRate_ftperyr, colour = river)) +
+ggplot(combined_data_clean, aes(x = After_date, y = AnnualShoalingRate_ftperyr, colour = river)) +
+  geom_point() + 
+  geom_line()+
+   labs(title = "Annual Shoaling Rate (ft/yr) By Pool",
+       x = "Survey Date",
+       y = "Annual Shoaling Rate (ft/yr)",
+       color = "River") +
+ # ylim(-30, 30)+
+  facet_wrap(~ pool)
+
+ggplot(combined_data_clean, aes(x = After_date, y = NetVolumeChange_CY, colour = river)) +
   geom_point() + 
@@
 
-ggplot(combined_data_year, aes(x = After_date, y = NetVolumeChange_CY, colour = river)) +
-  geom_point() + 
-  geom_line()+
-#  ylim(NA, 275)+
-  facet_wrap(~ river)
-
-ggplot(combined_data_year, aes(x = Before_date, y = NetVolumeChange_CY, color = river)) +
+ggplot(combined_data_clean, aes(x = Before_date, y = NetVolumeChange_CY, color = river)) +
   geom_point() +

### 2025-09-03 Draft_Cleaning.Rmd
catching up for capstone

@@
 library(progress)
+n_outer_folds = 5
 pb <- progress_bar$new(total = n_outer_folds)
@@
 n = nrow(gage_xgb_scaled)
-n_outer_folds = 5
 
@@
 
-gage_xgb_scaled_clean <- gage_xgb_scaled|>
-  select(-lead_7,-lead_14)
+gage_xgb_scaled_clean <- gage_xgb_scaled
 
@@
 # Optionally remove future-looking targets
-gage_scaled_clean <- gage_scaled_clean |> select(-lead_7, -lead_14)
+gage_scaled_clean <- gage_scaled_clean
 
@@
 # Optional: remove future-looking columns
-gage_scaled_clean <- gage_scaled_clean |> select(-lead_7, -lead_14)
+gage_scaled_clean <- gage_scaled_clean
 

### 2025-09-03 Lesson 6 examples--XGBoost.Rmd
catching up for capstone

---
title: "XGBoost"
author: "Abra Brisbin"
date: "11/1/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggformula)

library(ISLR) # for College data, used in hwk 4

library(caret)

library(xgboost)
library(DiagrammeR)

```

## Predicting Enrollment

```{r}

head(College)

```

```{r}

set.seed(123)
ctrl = trainControl(method = "cv", number = 5) # Increase this to 10 to get more consistent RMSEs
fit_coll = train(Enroll ~ ., 
             data = College,
             method = "xgbTree",
             tuneGrid = expand.grid(nrounds = 100, max_depth = 1:3, 
                                    eta = 0.3, gamma = 0, 
                                    colsample_bytree = .8, 
                                    min_child_weight = 1, subsample = 1),
             verbosity = 0, 
             trControl = ctrl)

fit_coll

```

### Variable importance

```{r}

varImp(fit_coll)

```

```{r}

import_matrix = xgb.importance(model = fit_coll$finalModel)
import_matrix
xgb.plot.importance(import_matrix)

```
## Double CV

```{r}

set.seed(123)
n = nrow(College)
n_outer_folds = 10

cv_pred = vector(length = n)

groups = rep(1:n_outer_folds, length = n)
cv_groups = sample(groups, n)

for(ii in 1:n_outer_folds){
  in_test = (cv_groups == ii)
  in_train = (cv_groups != ii)

  ctrl = trainControl(method = "cv", number = 5)
  fit_coll_DCV = train(Enroll ~ ., 
                       data = College[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = 100, max_depth = 1:3, 
                                              eta = 0.3, gamma = 0, 
                                              colsample_bytree = .8, 
                                              min_child_weight = 1, subsample = 1),
                       verbosity = 0, 
                       trControl = ctrl)

  # "Frankenstein" vector of predictions
  cv_pred[in_test] = predict(fit_coll_DCV, 
                             newdata = College[in_test, ])

}

# I'll use RMSE to assess the model, because that's the metric I used to select the model.
# The model with the best RMSE may or may not be the one with the best MAE.
# RMSE
sqrt(mean((cv_pred - College$Enroll)^2)) # This is the most honest 
                                         # assessment of accuracy

```

## Viewing Trees

Let's make a model with 2 trees, because that will be easier to view and make predictions about.
```{r}

set.seed(123)
ctrl = trainControl(method = "cv", number = 5)
fit_simple2 = train(Enroll ~ ., 
             data = College,
             method = "xgbTree",
             tuneGrid = expand.grid(nrounds = 2, max_depth = 1:2, 
                                    eta = 0.3, gamma = 0, 
                                    colsample_bytree = .8, 
                                    min_child_weight = 1, 
                                    subsample = 1),
             verbosity = 0, 
             trControl = ctrl)

```

```{r}

xgb.plot.tree(model = fit_simple2$finalModel, trees = 0)
xgb.plot.tree(model = fit_simple2$finalModel, trees = 1)
```

## Relating predictions to the trees
```{r}
# For a regression tree, the predictions are simply the sum of the Values in their leaves, + .5.

predictions2 = predict(fit_simple2)

head(predictions2)
College %>%
  dplyr::select(F.Undergrad, Accept) %>%
  head()
```
Abilene Christian University:  Tree 0 value = 100.426697
Tree 1 value = 202.066833
```{r}
100.426697 + 202.066833 +.5 # Matches the predicted value from predict()
```
### Classification trees

```{r}
set.seed(123)
ctrl = trainControl(method = "cv", number = 5)
private_simple2 = train(Private ~ ., 
             data = College,
             method = "xgbTree",
             tuneGrid = expand.grid(nrounds = 2, max_depth = 1:2, eta = 0.3, gamma = 0, colsample_bytree = .8, min_child_weight = 1, subsample = 1),
             trControl = ctrl)

private_simple2
```

```{r}
xgb.plot.tree(model = private_simple2$finalModel, trees = 0)
```

```{r}
xgb.plot.tree(model = private_simple2$finalModel, trees = 1)
```

```{r}
predictions_priv = predict(private_simple2, type = "prob")
head(predictions_priv)
College %>%
  dplyr::select(Outstate, Accept, Enroll, F.Undergrad, Room.Board) %>%
  head()
```
Abilene Christian University:  Tree 0: -0.156000003
Tree 1:  -0.412369579
```{r}
# For classification, the sum of the Values in the relevant leaves
# gives the log odds of Y = 1
a = 0 + exp(-0.156000003-0.412369579)
1/(1+a) # Prob(Private), matches the value from predict()
```

Y = 1 is "Yes", Private.
```{r}
contrasts(College$Private)
```

##################################

### 1 Tree
```{r}

set.seed(123)
ctrl = trainControl(method = "cv", number = 5)
fit_simple = train(Enroll ~ ., 
             data = College,
             method = "xgbTree",
             tuneGrid = expand.grid(nrounds = 1, max_depth = 1, 
                                    eta = 0.3, gamma = 0, 
                                    colsample_bytree = .8, 
                                    min_child_weight = 1, 
                                    subsample = 1),
             verbosity = 0, # suppresses a warning about 
                            # deprecated version of the 
                            # objective function
                            # (not necessary for classification)
             trControl = ctrl)

fit_simple

```

### 100 Trees

```{r}

#xgb.plot.tree(model = fit_coll$finalModel) # would show all trees
xgb.plot.tree(model = fit_coll$finalModel, trees = 0)
```

### Regression tree predictions
```{r}
xgb.plot.tree(model = fit_simple$finalModel)
```
```{r}
# baseline = 0.5, so each prediction is 0.5 + the Value of its leaf.
# darker arrows are the "yes" direction.
predictions = predict(fit_simple)
head(predictions)
head(College$Accept)
```

```{r}

xgb.plot.tree(model = fit_simple2$finalModel, trees = 0)

```

```{r}
xgb.plot.tree(model = fit_simple2$finalModel, trees = 1)
```

```{r}
# For a regression tree, the predictions are simply the sum of the Values in their leaves, + .5.

predictions2 = predict(fit_simple2)

head(predictions2)
College %>%
  select(F.Undergrad, Apps) %>%
  head()
```
Abilene Christian University:  Tree 0 value = 221.183502
Tree 1 value = 101.9478
```{r}
221.183502 + 101.9478 +.5 # Matches the predicted value from predict()
```

### Display the depth of the trees
```{r}
xgb.plot.deepness(model = fit_coll$finalModel)
```

# Classification

## 1 tree
```{r}
set.seed(123)
ctrl = trainControl(method = "cv", number = 5)
private_simple = train(Private ~ ., 
             data = College,
             method = "xgbTree",
             tuneGrid = expand.grid(nrounds = 1, max_depth = 1:2, eta = 0.3, gamma = 0, colsample_bytree = .8, min_child_weight = 1, subsample = 1),
             trControl = ctrl)

private_simple
```

```{r}
xgb.plot.tree(model = private_simple$finalModel)
```

```{r}
 # newdata argument is optional here, but seems necessary on the homework.

predictions_priv = predict(private_simple, 
                           newdata = College, 
                           type = "prob")

head(predictions_priv)
College %>%
  select(Outstate, Accept, Enroll) %>%
  head()
```

Abilene Christian University:  Value = -0.156000003
```{r}

a= 0 + exp(-0.156000003) # 0 is the log odds from the baseline (.5)
1/(1+a)
```

## Classification with 2 trees

### 2025-09-03 CSAT_Dredge_Data.R

@@ -106,4 +106,11 @@ write_csv(CSAT_dredge_data, "C:/workspace/DRAFT/CSAT_Data/Dredged_Date_Vol.csv")
 ggplot(CSAT_dredge_data, aes(x = SurveyDate, y = VolumeChange, colour = River)) +
   geom_point(size = 2)  # Scatter plot
 
+Miss<-dredge_data |>
+  filter(RIVER == "Mississippi_River")
+
+IL<-dredge_data |>
+  filter(RIVER == "Illinois_Waterway")
+  mean(IL$VOLUMEDREDGED)
+  mean(Miss$VOLUMEDREDGED)
 

### 2025-09-03 Kate_files_Summer_2024/Kate_files_Summer_2024/Sub-Reach Analysis/SubReach_Analysis.py

@@ -1,0 +1,1082 @@
+# -*- coding: utf-8 -*-
+"""
+initial code started 6/10/2024
+goal: analyze at sub-reach level
+
+6/14/2024: steps changed to include a way to determine which reaches are intersected using NCF feature service
+6/26/2024: step 1 completed
+7/12/2024: step 2 completed
+7/29/2024: step 3 completed
+8/5/2024: script validated
+"""
+import arcpy
+import requests
+import pandas as pd
+import geopandas as gpd
+import numpy as np
+from pathlib import Path
+import os
+import shutil
+import rasterio
+from rasterio.mask import mask
+from shapely.geometry import mapping
+from dateutil.relativedelta import relativedelta
+from datetime import datetime
+
+# Inputs that change per user
+PolygonFile = Path('..\\data\\shp\\MVR_AOI.shp')  # Path to the constraining shapefile
+ParentDir = Path('..\\output\\CEMVR\\19990101_to_20241231_omittedtable')  # Path to the CSAT outputs
+# The CSAT outputs should be located in the output folder, right? Should this be default?
+
+# Number of years for analysis
+NY = 27
+
+# Name of AOI defined as the name of the shapefile
+aoi_name = PolygonFile.stem
+
+# Output directories: the paths below bank on this script being placed in the "source code" folder within CSAT
+BaseOutputDir = Path('..\\output\\Sub-Reach Analysis')
+OutputDir = os.path.join(BaseOutputDir, aoi_name)  # Folder for the outputs of this code
+MovingDir = os.path.join(OutputDir, "WorkingFiles")  # Created path for the intersecting reaches
+
+if os.path.exists(OutputDir):
+    response = input(f"Directory for {aoi_name} exists. Overwrite? (y/n): ")
+    if response.lower() == 'y':
+        shutil.rmtree(OutputDir)
+    else:
+        print("Operation cancelled.")
+        exit()
+os.makedirs(MovingDir)
+# Do we want to keep the above lines in the code?
+
+
+
+# variables from feature service and feature service url for channel reaches
+feature_service_url = "https://services7.arcgis.com/n1YM8pTrFmm7L4hs/ArcGIS/rest/services/National_Channel_Framework/FeatureServer/2/query"
+fields = "channelreachidpk,depthauthorized,channelareaidfk,sdsfeaturename"
+arcpy.env.overwriteOutput = True
+
+
+
+"Step 1: Access feature service and determine which reaches intersect with the AOI. These reaches are then pulled from " \
[…]

### 2025-09-03 model.py

@@ -1,0 +1,61 @@
+# model.py
+
+import tensorflow as tf  # Keep this
+# Remove or comment out these lines:
+# from tensorflow import keras
+# from tensorflow.keras.models import Sequential
+# from tensorflow.keras.layers import LSTM, Dense, Dropout, Activation
+
+from config import (
+    INPUT_SEQUENCE_LENGTH, OUTPUT_SEQUENCE_LENGTH,
+    LSTM_UNITS_1, LSTM_UNITS_2, LSTM_DROPOUT_1, LSTM_DROPOUT_2,
+    LSTM_ACTIVATION, DENSE_ACTIVATION, FINAL_ACTIVATION,
+    OPTIMIZER_LR, N_FEATURES_TO_KEEP
+)
+
+def build_lstm_model(n_features: int) -> tf.keras.models.Sequential: # Use tf.keras... here
+    """Builds the LSTM model architecture."""
+    print("Building LSTM model...")
+    if n_features <= 0:
+        raise ValueError("Number of features must be positive.")
+
+    opt = tf.keras.optimizers.Adam(learning_rate=OPTIMIZER_LR) # Use tf.keras...
+
+    model = tf.keras.models.Sequential(name=f"LSTM_{INPUT_SEQUENCE_LENGTH}in_{OUTPUT_SEQUENCE_LENGTH}out") # Use tf.keras...
+    model.add(tf.keras.layers.LSTM( # Use tf.keras...
+        LSTM_UNITS_1,
+        activation=LSTM_ACTIVATION,
+        return_sequences=True, # Return sequences for the next LSTM layer
+        input_shape=(INPUT_SEQUENCE_LENGTH, n_features),
+        name='lstm_1'
+    ))
+    model.add(tf.keras.layers.Dropout(LSTM_DROPOUT_1, name='dropout_1')) # Use tf.keras...
+
+    # Add second LSTM layer (optional, based on config/analysis)
+    if LSTM_UNITS_2 > 0:
+        model.add(tf.keras.layers.LSTM( # Use tf.keras...
+            LSTM_UNITS_2,
+            activation=LSTM_ACTIVATION,
+            return_sequences=False, # Only return the last output for the Dense layer
+            name='lstm_2'
+        ))
+        model.add(tf.keras.layers.Dropout(LSTM_DROPOUT_2, name='dropout_2')) # Use tf.keras...
+    # Else: If only one LSTM layer, ensure its return_sequences=False
+
+    # Output layer
[…]
